Noncanonical roles of ATG5 and membrane atg8ylation in retromer assembly and function

Version 1

Abstract

Abstract

ATG5 is one of the core autophagy proteins with additional functions such as noncanonical membrane atg8ylation, which among a growing number of biological outputs includes control of tuberculosis in animal models. Here we show that ATG5 associates with retromer’s core components VPS26, VPS29 and VPS35 and modulates retromer function. Knockout of ATG5 blocked trafficking of a key glucose transporter sorted by the retromer, GLUT1, to the plasma membrane. Knockouts of other genes essential for membrane atg8ylation, of which ATG5 is a component, affected GLUT1 sorting, indicating that membrane atg8ylation as a process affects retromer function and endosomal sorting. The contribution of membrane atg8ylation to retromer function in GLUT1 sorting was independent of canonical autophagy. These findings expand the scope of membrane atg8ylation to specific sorting processes in the cell dependent on the retromer and its known interactors.

Introduction

The canonical autophagy pathway, ubiquitous in eukaryotes, is manifested by the emergence of intracellular membranous organelles termed autophagosomes that capture cytoplasmic cargo destined for degradation in lysosomes 1. Progress has been made in understanding molecular mechanisms governing canonical autophagosome biogenesis in mammalian cells 2 including ATG9A-vesicles 3–7 and membranous prophagophores 8 as precursors to LC3-positive phagophores, their reshaping as cups 9 and dynamic interactions with the omegasomes 10,11, phagophore expansion through lipid transfer 12–14, cargo recognition and sequestration 15,16, closure of the phagophore to form a double membrane phagosomes 17–20, and fusion of autophagosomes with endosomal and lysosomal organelles leading to cargo degradation 21 or secretion 22.

However, and contemporaneously with these major advances in understanding canonical autophagy, it has become evident that autophagy genes (ATGs) 23 have additional functions in mammalian cells that do not fit the canonical model 24,25. Among such noncanonical manifestations are LAP (LC3-associated phagocytosis) 26, LANDO (LC3-associated endocytosis) 27, LC3-associated micropinocytosis (LAM) 28, conjugation of ATG8 to single membranes (CASM) 29,30, v-ATPase-ATG16L1-induced LC3 lipidation (VAIL) 31–33, ER-phagy mediated by the V-ATPase-ATG16L1-LC3C axis (EVAC) 34, LyHYP (lysosomal hypersensitivity phenotype) 35, membrane damage repair 36–39, and two distinct forms of secretory autophagy 22, SALI (secretory autophagy during lysosome inhibition) 40 and LDELS (LC3-dependent EV loading and secretion) 41. These processes depend on or are associated with the phospholipid conjugation cascade of mammalian ATG8 proteins (mATG8s) 42 and collectively (including the canonical autophagy) represent diverse manifestations of membrane atg8ylation as a broad membrane stress, damage and remodeling response 25.

The factors governing membrane atg8ylation include two enzymatic cascades with the ATG12-ATG5 and mATG8-phosphatidylethanolamine (PE) covalent conjugates as their products. The ATG12-ATG5 conjugate 42 combines with ATG16L1 43 or TECPR1 to form E3 ligases 38,39,44 to direct mATG8-PE conjugation and atg8ylation of target membranes. All known E3 enzymes contain the ATG12-ATG5 conjugate 38,39,42,44. To form this conjugate, the ubiquitin like molecule ATG12 is activated by ATP, and transferred via E1 (ATG7) and E2 (ATG10) to ATG5. In preparation for the next step, an ATG12-ATG5 containing E3 ligase activates its substrate mATG8-ATG3 to transfer ATG8 and form an amide bond with aminophospholipids 42. However, there are additional branches of these conjugation cascades, whereby ATG12 can make a noncanonical sidestep conjugate with ATG3 (ATG12-ATG3) 41,45 which is enhanced in the absence of ATG5 35. Apart from its role in atg8ylation, ATG5 35,46 and potentially other ATG genes 47–49 have autophagy-independent functions. Many of the indications that such functions exist come from in vivo studies 50–52. For example, in the case of murine models of Mycobacterium tuberculosis infection, atg8ylation machinery protects against tuberculosis pathogenesis but Atg5 knockout has a particularly strong phenotype exceeding other atg8ylation genes, indicating that ATG5 possesses atg8ylation (mATG8s lipid conjugation) independent functions 35,46,53–56. As recently reported, one such function involves LyHYP contributing to enhanced exocytosis specifically in cells devoid of ATG5 but not of other ATG genes 35. These developments indicate that ATG5 plays noncanonical roles in autophagy-independent vesicular transport events and homeostasis of the endolysosomal system.

Within the endosomal network, the mammalian retromer complex 57 is one of the several systems regulating vectoral transport of membranes and proteins. These systems include, among others, HOPS, CORVET, retriever, CCC/commander, and retromer 57–64 complexes that often contain and sometimes share VPS subunits. Retromer is a heterotrimer of VPS26A/B, VPS29 and VPS35 subunits 65, first identified in yeast 66. Retromer sorts an array of endosomal cargo in cooperation with cognate sorting nexins (SNX) 67. Specifically, this includes SNX-BAR proteins SNX1/2-SNX5/6 68,69, the mammalian paralogs of yeast Vps5p and Vps17p 66 which can function autonomously as an ESCPE-I complex 70, SNX-PX protein SNX3 71, and SNX-FERM protein SNX27 containing PX, PDZ and FERM domains 72 and sorting various transporter proteins and signaling receptors including the glucose transporter GLUT1 (SLC2A1) 73. Retromer, together with adaptors, contributes to the complex protein sorting within the endosomal system 74–76. Here, using unbiased proteomic approaches and follow-up mechanistic analyses, we report that retromer complex is among ATG5 interactors. We show that membrane atg8ylation, of which ATG5 is an essential component, affects retromer-dependent cargo sorting. This function is independent of the canonical autophagy pathway or a well-studied form of noncanonical membrane atg8ylation, CASM. Our findings offer a paradigm shift connecting ATG5 and membrane atg8ylation with the retromer system and its function in the endosomal cargo sorting, expanding the scope of atg8ylation machinery and its special component ATG5 beyond the current models.

Results

ATG5 associates with the retromer complex

The unique aspects of Atg5 (Fig. 1A), outside of its role in canonical autophagy, have strong ex vivo and important in vivo phenotypes 52. Inactivation of Atg5 in myeloid lineage renders mice excessively susceptible to acute infection with Mycobacterium tuberculosis attributed to excessive inflammation 35,46,53–56, which did not extend to other phases of infection as tested here in a murine model of latent tuberculosis 77–79 (Fig. S1A-D). In a previous study, we could only partially explain the cellular parameters associated with acute infection, which included degranulation in neutrophils from Atg5fl/fl LysM-Cre+ mice 35. Furthermore, markers of LyHYP (lysosome hypersensitivity phenotype) 35 in ATG5 knockout (KO) cells, such as galectin 3 (Gal3) response, could not be explained by the exocytic phenomena 35. We thus sought to uncover additional processes (Fig. 1A) affected by ATG5 at the intracellular level. To include factors beyond the known processes, we analyzed proteomic data obtained with APEX2-ATG5 35 and compared APEX2-ATG5WT with APEX2-ATG5K130R (an ATG5 mutant deficient in conjugation to ATG12) in cells treated with LLOMe, an agent routinely used to cause lysosomal damage 80–85 (Figs. 1B and S2A,B). The proximity biotinylation (APEX2)-based proteomic data indicated the presence in the vicinity of APEX2-ATG5 of several protein complexes that control important membrane trafficking pathways including retromer (Fig. 1B). Beside the retromer VPS subunits, there were very few other VPS proteins (Fig. S2C). VPS proteins were observed in ATG5 proteomic data by others 86 (Fig. S2D). VPS26 and VPS29 showed statistically significant increase in proximity to ATG5 in cells subjected to LLOMe treatment (Fig. S2A-C). Using APEX2-SGALS1 proximity biotinylation LC-MS/MS as a control, we did not observe enrichment of retromer subunits in cells treated with LLOMe (Fig. S2E).

ATG5 interacts with retromer.

A. ATG5 functions. X, a postulated additional function. B. 2D scatter plot (log2 fold changes; color coded p value cutoff matrix for comparisons between samples as per the lookup table) of proximity biotinylation LC/MS/MS datasets: FlpIn-HeLaAPEX2-ATG5-WT cells (X-axis) and FlpIn-HeLaAPEX2-ATG5-K130R (Y-axis) treated with 2 mM LLOMe for 30 min ratioed vs. HeLaAPEX2-ATG5-WT without LLOMe treatment (Ctrl). C-E. Co-IP analyses and quantification of VPS26A (C), VPS29 (D) VPS35, (E), interaction with ATG5 in HeLa cells treated with or without 2 mM LLOMe for 30 min. Data, means ± SE (n=3); unpaired t-test; p values indicated above the bars.

Association of endogenous ATG5 with endogenous retromer subunits (VPS26A, VPS29 and VPS35) was tested in coimmunoprecipitation (co-IP) experiments (Fig. 1C-E). The interaction between retromer and ATG5 was detected in these experiments and showed increased association in cells subjected to lysosomal damage with LLOMe (Fig. 1C-E). Thus, ATG5, a key component of the known atg8ylation E3 ligases 42,87, is found in protein complexes with the retromer and this association is enhanced upon lysosomal damage.

Retromer affects a subset of responses to lysosomal damage

During lysosomal damage a number of processes are set in motion to repair, remove and regenerate/replenish lysosomes 37,80,88–91 including membrane atg8ylation-dependent processes of repair by ESCRT machinery 92 and by lipid transfer via ATG2 85,93. We have reported that ATG5 KO cells display elevated Gal3 response to lysosomal damage agents including LLOMe 35. Gal3 is one of the well-characterized sentinel galectins alerting cellular homeostatic systems to lysosomal damage 37,80,81,88,94. Based on the observed interactions between ATG5 and retromer, we tested whether retromer, like ATG5 35, affected Gal3 response to lysosomal damage. We used the previously established quantitative high content microscopy (HCM) approach, which provides unbiased machine-driven image acquisition and data analysis 8,36,80,88,94. In the experiments herein HCM was based on >500 valid primary objects/cells per well, with a minimum of 5 wells per sample (the independent biological replicates being n≥3) as described 35. Upon treatment with LLOMe, HeLaVPS35-KO cells had increased Gal3 puncta relative to HeLaWT cells, comparable to HeLaATG5-KO (Fig. 2A,B) 35. The Gal3 phenotype in VPS35 knockout cells was complemented by expression of GFP-VPS35 (Fig. 2C,D). The observed elevated Gal3 recruitment to damaged lysosomes in the absence of VPS35 cannot be explained by increased pools of LAMP2 organelles in the cells, since the overall complement of lysosomes (quantified by LAMP2 puncta/cell) did not increase in LLOMe treated HeLa cells (Fig. S3A,B). Another marker of lysosomal damage, ubiquitination response 95, was elevated in both VPS35 and ATG5 deficient cells (Figs. 2E,F). This phenotype was confirmed in another cell line, Huh7 (Fig. S3C,D). Thus, VPS35 and ATG5 defects have a similar effect on lysosomes, previously characterized and termed lysosomal hypersensitivity phenotype/LyHYP 35.

Retromer affects lysosomal sensitivity to damage.

A,B. High content microscopy (HCM) imaging and quantification of Gal3 response (puncta/cell of endogenous Gal3 profiles stained for immunofluorescence) in HeLaWT, HeLaATG5-KO, and HeLaVPS35-KO cells subjected to lysosomal damage by Leu-Leu-O-Me ester hydrobromide (LLOMe; 2mM, 30 min). HCM, an unbiased machine-driven image acquisition and data analysis based on presets of >500 valid primary objects/cells per well (representative images shown; white mask, cell; red masks, Gal3 puncta), with a minimum of 5 wells per sample (sampling error), and n≥3, independent biological replicates (experimental error) in separate 96-well plates. Data, means ± SE (n=3), two-way ANOVA with Tukey’s multiple comparisons. C,D. Complementation analysis of VPS35KO LyHYP (lysosome hypersensitivity) phenotype monitored by HCM quantification of Gal3 puncta in HeLaVPS35-KO cells transfected with GFP (control) or GFP-VPS35 expressing plasmids. Data, means ± SE (n=3); one-way ANOVA with Tukey’s multiple comparisons. E,F. Comparative HCM analysis of ubiquitin (immunofluorescence; FK2 antibody) response to lysosomal damage (LLOMe; 2mM, 30 min) in HeLaATG5-KO and HeLaVPS35-KO cells. Yellow profiles, colocalization of ubiquitin and LAMP1. Top panels, ubiquitin immunostaining alone. Data, means ± SE (n=4), two-way ANOVA with Tukey’s multiple comparisons. G,H. HCM quantification of ALIX localization to endolysosomal compartments (% of LAMP1 profiles positive for ALIX immunostaining) in Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells following lysosomal damage. Yellow profiles, colocalization of ALIX and LAMP1. Data, means ± SE (n=3); two-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line/condition.

Not all aspects of LyHYP, previously observed in ATG5 knockouts 35, were seen in VPS35 KO cells. In the case of ATG5 inactivation, the recruitment of ALIX, an ESCRT component involved in early stages of lysosomal repair 80,96,97 is abrogated 35. In ATG5 KO cells (Huh7ATG5-KO), recruitment of ALIX to lysosomes was prevented, whereas in VPS35 KO cells (Huh7VPS35-KO and HeLaVPS35-KO) ALIX was efficiently recruited to lysosomes upon LLOMe treatment (Figs. 2G,H and S3C,D).

The elevated Gal3 and ubiquitin markers of increased lysosomal damage in cells with inactivated VPS35 or ATG5 suggest that both may play a related role in lysosomal resilience to membrane damage, which complements the previously observed effects of aberrant VPS35/retromer on lysosomal morphology and proteolytic capacity 98,99. The dissimilarities between ATG5 and retromer effects on the ALIX recruitment component can be explained by the previously described perturbed function of the ATG conjugation machinery specifically in ATG5 KO cells affirming the unique additional roles of ATG5 35. Nevertheless, the striking similarities between ATG5 KO and VPS35 KO effects on heightened Gal3 and ubiquitin responses to lysosomal membrane damage suggest that the two systems (membrane atg8ylation and retromer) closely intersect.

ATG5 affects retromer function

To avoid the confounding, and possibly indirect, effects of lysosomal damage on endolysosomal sorting and trafficking in cells treated with LLOMe, it was necessary to test the effects of ATG5 KO on retromer function under basal conditions with unperturbed lysosomes. Our co-IP analyses (Fig. 1C-E) indicated that ATG5 can be found in protein complexes with retromer subunits even without the lysosomal damage. We further confirmed this in co-IPs showing that GFP-VPS35 and YFP-VPS29 were found in complexes with endogenous ATG5 in resting cells, whereas an isotype IgG control did not pulldown retromer components (Fig. 3A). Reverse co-IPs with GFP-VPS35 and YFP-VPS29 detected endogenous ATG5 and ATG12-ATG5 complexes in GFP/YFP immunoprecipitates (Fig. 3B).

ATG5 affects sorting of the retromer cargo GLUT1.

A. Co-IP analysis of endogenous ATG5 with GFP-VPS35 or YFP-VPS29 (transient transfection). B. Reverse Co-IP analysis GFP-VPS35 or YFP-VPS29 (transient transfection) and endogenous ATG5. C. Confocal images illustrating localization of GLUT1 and LAMP2 in Huh7WT, Huh7VPS35-KO, and Huh7ATG5-KO cells. Scale bar, 10 μm. D,E. HCM quantification of GLUT1 (endogenous protein immunostaining) puncta/cell (D) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (E) in Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. F,G. HCM quantification of GLUT1-SNX27 overlap (% of SNX7 area positive for GLUT1) in Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. H,I. Immunoblot analysis and quantification of proteins in lysosomes purified/enriched by LysoIP (immunoisolation with TMEM192-3xHA) from Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. TMEM192-2xFLAG, negative control. Data, means ± SE (n=3), one-way ANOVA with Tukey’s multiple comparisons. HCM images in panel F, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

Having established that ATG5 and retromer interact under basal conditions, we tested the effects of ATG5 KO on localization of a well-established cargo for retromer-dependent sorting, GLUT1 73,100. We used Huh7 cells, where we generated a panel of membrane atg8ylation and VPS35 CRISPR knockouts (Fig. S4A), because they have a well-defined localization of GLUT1 at the plasma membrane or as intracellular profiles (Fig. 3C), fully amenable (and comparatively more reliable than in HeLa cells) to unbiased quantification by HCM. ATG5 KO in Huh7 cells resulted in redistribution of GLUT1 from the plasma membrane to intracellular punctate profiles (Figs. 3D and S4B). This phenotype was identical to the one observed in VPS35 KO (Fig. 3D and S4B). The increased intracellular GLUT1 puncta overlapped with the lysosomal marker LAMP2 in both ATG5 KO and VPS35 KO cells (Figs. 3C,E and S4C). A similar albeit morphologically less distinct phenotype was observed in HeLa cells (Figs. S4D-H). GLUT1 is sorted by retromer via VPS26’s interactor SNX27, with GLUT1 (and similar cargo) being captured by the PDZ domain of SNX27 72,73. In WT cells, SNX27 was found in abundant intracellular profiles (Fig. 3F). In both ATG5 KO and VPS35 KO cells, GLUT1 and SNX27 relocalized to the same intracellular compartment (Fig. 3F,G). We carried out lysosomal purification using the well-established LysoIP method 80,91,94,101, and quantified levels of GLUT1 and SNX27 in lysosomal preparations that were positive for LAMP2 and devoid of GM130 (Golgi) and PDI (ER) (Fig. 3H). Both GLUT1 and SNX27 were enriched in lysosomal preparations from ATG5 KO and VPS35 KO cells relative to the WT cells (Fig. 3H). These experiments biochemically identify the lysosomes as a compartment to which GLUT1 and SNX27 relocalize in ATG5 KO and VSP35 KO cells. The above findings show that a loss of ATG5 affects retromer-controlled trafficking events.

Membrane atg8ylation apparatus contributes to retromer function in GLUT1 sorting

ATG5 is a part of the membrane atg8ylation machinery 25,42,87, also known under the term ‘LC3 lipidation’ 102, which functions within the canonical autophagy pathway but is also engaged in a wide array of noncanonical processes 25,38,39,44,103,104. We tested whether other known components of the atg8ylation apparatus affect retromer. The KOs in Huh7 cells of the components of the prototypical atg8ylation E3 ligase 25,42 included: ATG5 and ATG16L1, as well as the previously characterized Huh7 lines knocked out for E1 and E2 enzymes, ATG7 35 and ATG3 37. These KOs all caused entrapment of GLUT1 in intracellular punctate profiles (Figs. 4A, B) as well as an increase in GLUT1 colocalization with lysosomes (GLUT1+ LAMP2+ profiles; Fig. 4C,D). We tested whether retromer component levels were altered in atg8ylation mutants and found no change in total cellular levels of VPS26, VPS29 and VPS35, tested in ATG3, ATG5 and ATG7 KO Huh7 cells (Fig. S5A,B) as well as in ATG5 KO HeLa cells (Fig. S5C,D).

Membrane atg8ylation apparatus affects sorting of the retromer cargo GLUT1.

A,B. HCM quantification of GLUT1 (endogenous protein immunostaining) puncta/cell in Huh7WT, Huh7ATG3-KO, Huh7ATG5-KO, Huh7ATG7-KO, Huh7ATG16-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. C,D. HCM quantification of GLUT1 (endogenous protein immunostaining) colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) in Huh7WT, Huh7ATG3-KO, Huh7ATG5-KO, Huh7ATG7-KO, Huh7ATG16-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. E,F. HCM quantification of GLUT1 (endogenous protein immunostaining) colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) in HeLaWT, and HeLaHexa-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

Furthermore, comparison of the previously characterized HeLaHexa-KO cells 105, with all principal mATG8s inactivated, with their isogenic parental HeLaWT cells, showed increased GLUT1 on lysosomes in HeLaHexa-KO (GLUT1+ LAMP2+ profiles; Fig. 4E,F). We observed a similar phenotype in separate triple knockouts (TKOs) of LC3 subfamily and GABARAP subfamily of mATG8s 105 compared to HeLaHexa-KO (Fig. S5B). Thus, the entire membrane atg8ylation machinery was engaged in and required for proper GLUT-1 sorting.

Canonical autophagy cannot explain effects of membrane atg8ylation on GLUT1 sorting

Membrane atg8ylation and mATGs play roles in diverse processes 25 including canonical autophagy 1. We tested whether the participation of atg8ylation in canonical autophagy, previously reported to affect retromer-dependent cargo sorting under nutrient-limiting conditions 106,107, is the reason for reduced trafficking of GLUT1 in our experiments. We used KOs in Huh7 cells of ATG13 and FIP200/RB1CC1, two obligatory components of the canonical autophagy initiation machinery 1, to test whether canonical autophagy under the basal conditions used in our study was responsible for the effects on GLUT1 sorting. We found that in the Huh7 cells knocked out for ATG13 35 or FIP200 37 there was no increase in intracellular GLUT1 puncta in contrast to the increase observed with the isogenic ATG5 and VPS35 mutants (Figs. 5A,B). This was mirrored by no increase of GLUT1+LAMP2+ profiles in ATG13 and FIP200 KO cells whereas colocalization between GLUT1 and LAMP2A increased in ATG5 and VPS35 mutants (Fig. 5BC). To further confirm that autophagy initiation mutant cells retained the ability to perturb GLUT1 trafficking due to defective atg8ylation, we knocked down ATG5 in FIP200 KO cells (Fig. S5D) and found that GLUT1 puncta and GLUT1+LAMP2+ profiles increased even in the FIP200 KO background to an extend nearing in strength the effects of VPS35 knockout (Figs. 5D-F and S5C). Thus, we conclude that atg8ylation but not canonical autophagy affects retromer function under basal conditions and that a functional atg8ylation apparatus is required for proper sorting of the retromer cargo GLUT1.

Canonical autophagy does not affect sorting of the retromer cargo GLUT1.

A,B. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell in in Huh7WT, Huh7ATG13-KO, Huh7FIP200-KO, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. C,D. HCM quantification of GLUT1 (immunostaining of endogenous protein) colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) in in Huh7WT, Huh7ATG13-KO, Huh7FIP200-KO, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. E-G. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell (E) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (F) in Huh7WT, Huh7FIP200-KO, Huh7FIP200-KO + siATG5, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

ATG5 and membrane atg8ylation machinery affect Rab7 localization

Rab7 is considered to be an interactor of the retromer 108. Rab7 translocates in cells lacking the retromer component VPS35 to lysosomes 109 including endolysosomal domains with lysosomaly positioned mTORC1 regulatory machinery 110. We thus tested whether absence of ATG5 and components of the atg8ylation apparatus affected Rab7 localization. ATG5, VPS35 as well as knockouts of all major components of the atg8ylation in Huh7 cells, displayed increased Rab7+ cytoplasmic profiles (Fig. 6A,B) and overlap between Rab7 and lysosomes (LAMP2) (Figs. 6C and S6A). This was confirmed biochemically using lysosomal purification by LysoIP and immunoblotting for Rab7: there was an increase of Rab7 in lysosomal preparations from ATG5 and VPS35 KO cells (Fig. 6D,E).

Membrane atg8ylation machinery is required for proper RAB7 localization.

A-C HCM quantification of Rab7 (endogenous protein immunostaining) puncta/cell (B) and Rab7 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (C) in Huh7WT, Huh7ATG3-KO, Huh7ATG5-KO, Huh7ATG7-KO, Huh7ATG16-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. D-F. HCM quantification of Rab7 (endogenous protein immunostaining) puncta/cell (E) and Rab7 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (F) in Huh7WT, Huh7ATG13-KO, Huh7FIP200-KO, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. G,H. Immunoblot analysis and quantification of proteins in lysosomes purified/enriched by LysoIP (immunoisolation with TMEM192-3xHA) from Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. TMEM192-2xFLAG, negative control. Data, means ± SE (n=3), one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

This effect extended to other components of the atg8ylation machinery, as KOs in ATG3, ATG7 and ATG16L1 caused a similar effect (Figs. 6A-C and S6A). In contrast, KOs in genes specific for canonical autophagy (ATG13 and FIP200) did not alter intracellular distribution of Rab7 (Figs. 6F-H and S6B). Rab7 distribution was affected by ATG5 knockdown (Fig. S5D) in FIP200 KO cells (Figs. S6B-E). These experiments mirror the effects of ATG5 and atg8ylation on GLUT1 trafficking, showing that ATG5 and atg8ylation machinery but not canonical autophagy is required for proper localization of Rab7.

Agonists of endolysosomal atg8ylation process CASM affect GLUT1 sorting

Membrane atg8ylation has multiple presentations that include not only canonical autophagy but also encompass processes such as CASM 103. We thus asked whether induction of CASM 93 could affect GLUT1 trafficking. Using LLOMe as one of the inducers of CASM 93, we detect increased membrane atg8ylation (LC3 puncta formation) in response to lysosomal damage (Fig. 7A,B). When cells were treated with LLOMe 93 and another CASM inducer, monensin 29,111,112, GLUT1 trafficking to the plasma membrane was negatively affected and instead GLUT1 accumulated intracellularly and trafficked to lysosomes (Figs. 7C-E and S7A). Thus, lysosomal perturbation and induction of CASM may affect retromer-dependent sorting of GLUT1.

CASM induction affects sorting of the retromer cargo GLUT1.

A,B. HCM imaging and quantification of LC3 response (puncta/cell of endogenous LC3 immunofluorescent profiles) in HeLaWT, and HeLaVPS35-KO cells in response to lysosomal damage by LLOMe (1mM, 30 and 60 min). Scale bar, 20 μm. Data, means ± SE (n=5), one-way ANOVA with Tukey’s multiple comparisons. C-E. HCM quantification of GLUT1 response (puncta/cell of endogenous GLUT1, D) and its localization to endolysosomal compartments (% of LAMP1 profiles positive for GLUT1 immunostaining, E) in Huh7WT treated with or without Monensin (100µM), LLOMe (100µM), and Bafilomycin A1 (100nM) for 45 minutes. Scale bar, 20 μm. Data, means ± SE (n=5), one-way ANOVA with Tukey’s multiple comparisons. F-H. Analysis of GLUT1 puncta/cell (F) and its localization to endolysosomal compartments (% of LAMP1 profiles positive for GLUT1 immunostaining, G) phenotype monitored by HCM quantification in Huh7WT cells transfected with GFP (control) or GFP-Rab7WT, GFP-Rab7Q67L, and GFP-Rab7T22N, expressing plasmids. Scale bar, 20 μm. Data, means ± SE (n=3); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line/condition.

It was previously reported that under glucose starvation conditions, LC3A recruits TBC1D5, a GAP for Rab7 113–115, away from this small GTPase 106 and a similar phenomenon could be occurring during CASM. However, under CASM-inducing conditions, no change in interactions between TBC1D5 and LC3A were detected (Fig. S7B,C) suggesting that CASM does not affect the status of the TBC1D5-Rab7 system. Moreover, expression of constitutively active Rab7Q67L did not promote trafficking of GLUT1 to plasma membrane but rather caused accumulation of intracellular GLUT1 puncta (Fig. 7F-H).

Membrane atg8ylation and endolysosomal homeostasis affect GLUT1 sorting

A complication of interpreting CASM as a membrane atg8ylation process responsible for effects on retromer-dependent sorting is that CASM is elicited by lysosomal perturbations using lysosomal pH modifying agents 29,103,111,112, and lysosomal stress could be an independent factor preventing proper sorting of GLUT1. Endolysosomal membrane damage (LLOMe) or perturbations of luminal pH (monensin, bafilomycin A1) negatively affected GLUT1 sorting and caused it to accumulate in lysosomes (Fig. 7C-E). Since bafilomycin A1 does not induce CASM 93 but disturbs luminal pH, we conclude that it is the less acidic luminal pH of the endolysosmal organelles that is sufficient to interfere with the proper sorting of GLUT1.

The above experiments indicate the role of acidification and integrity of endolysosomal compartments in GLUT1 sorting. Membrane atg8ylation is important for lysosomal repair and homeostasis with specific downstream mechanisms. These processes include: (i) ATG2A, which transfers lipids during lysosomal repair 85 and is engaged on damaged lysosomes upon membrane atg8ylation by LC3A 93, and (ii) ESCRT-based lysosomal membrane repair 80,96,97,116, with membrane atg8ylation specifically recruiting an ESCRT regulator ALG2 92 and ESCRT-I component VPS37A 20. Hence, we tested whether these established systems, involved in maintenance of lysosomal membrane integrity, represented by ATG2 and VPS37, can explain the mechanism underlying the observed effects of membrane atg8ylation on retromer-dependent trafficking. When U2OS ATG2A/ATG2B double KO cells were compared to parental U2OS WT cells, we observed aberrant GLUT1 trafficking, presented as accumulation of cytoplasmic GLUT1 puncta and increased colocalization of GLUT1 with lysosomes (LAMP2A) (Figs. 8A-C and S8A). A similar effect was observed when the previously characterized VPS37A knockout cells 20 were tested (Figs. 8D-F and S8B). These effects were detected with or without induced lysosomal damage (Fig. 8A-F). We interpret these data as evidence that endolysosomal homeostasis and functionality, which is maintained by membrane atg8ylation at all times even under basal conditions, contributes to proper function of retromer-dependent sorting of its cognate cargo such as GLUT1 (Fig. 8E).

Membrane atg8ylation and endolysosomal homeostasis affect retromer-dependent sorting.

A-C. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell (B) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area, C) in U2OSWT and U2OSATG2A/B-DKO cells treated with or without LLOMe (100µM) for 45 minutes. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. D-F. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell (E) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area, F) in Huh7WT cand Huh7VPS37A-KO cells treated with or without LLOMe (100µM) for 45 minutes. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line/condition. G. Schematic: Membrane atg8ylation maintains membrane homeostasis under basal or stress conditions and, independently of canonical autophagy, affects retromer function under. A functional atg8ylation apparatus is required for proper sorting of the retromer cargo GLUT1.

ATG5 effects on retromer function exceed those of membrane atg8ylation

Finally, we addressed the strength of effects on GLUT1 trafficking observed with ATG5 KO vs. KOs in other membrane atg8ylation genes, which were significant in all experiments but presented effects lower in magnitude than ATG5 KO (Figs. 4 and 6). One possibility was that since ATG3, ATG5, and ATG7 all participate in the same process of membrane atg8ylation, that our KO mutants had unequal gene inactivation levels. This was ruled out by comparing effects of ATG3 KO, ATG5 KO and ATG7 KO cells for their effects on LC3 puncta formation in a well-defined system of starvation-induced autophagy. The results of these analyses indicated that all three KOs had indistinguishable effects on LC3 puncta formation during canonical autophagy induction by starvation (Fig. S8C,D). Thus, the stronger effects of ATG5 KO on retromer vis-à-vis KOs in other atg8ylation genes reflected additional action of ATG5. In the absence of ATG5, less VPS26 and VPS35 could be pulled down in co-IPs with YFP-VPS29 (Fig. S8E-H). We conclude that ATG5, in addition to contributing to the retromer-dependent sorting via membrane atg8ylation and its downstream effector mechanisms which maintain healthy endolysosomal organelles, has additional effects on the retromer.

Discussion

In this study, we have uncovered a new role of ATG5 and membrane atg8ylation machinery beyond the conventional function in canonical autophagy and the hitherto recognized noncanonical processes. ATG5 interacts with the retromer complex, whereas ATG5 and membrane atg8ylation affect retromer function in sorting of its cognate cargo GLUT1. Inactivation of ATG5 and of other membrane atg8ylation genes but not of the genes involved in canonical autophagy perturbs retromer-dependent trafficking. ATG5 and membrane atg8ylation display dual and combined effects on retromer function. The first one is through endolysosomal membrane maintenance ensuring proper cargo sorting by the retromer complex and its adaptor SNX27. The second one is via an association of ATG5 with the retromer complex. These and other activities of ATG5 underly the unique and often perplexing phenotypic manifestations of ATG5 observed in cells and in murine infection models 35,46,51,53–56, which served as the initial impetus for the present study. More generally, the finding that membrane atg8ylation influences retromer function further expands the range of this process specializing in homeostatic responses to membrane stress, damage and remodeling signals 25,104,117.

Mirroring the role of membrane atg8ylation on retromer function, a loss of VPS35 brings about LyHYP and influences cellular responses to lysosomal damage similarly to the inactivation of ATG5. The parallel between retromer and ATG5 is not perfect, as VPS35 knockout does not divert membrane repair protein ALIX from damaged lysosomes. This can be explained by sequestration of ALIX by the alternative conjugation complex ATG12-ATG3 formed in the absence of ATG5 35. The latter phenomenon renders ALIX unavailable for lysosomal repair 35. VPS35 knockout has no effects on ALIX, and hence the differences between ATG5 and VPS35 in the ALIX component of LyHYP. Nevertheless, inactivation of either ATG5 or VPS35 confers similarly heightened Gal3 80,81 and ubiquitin 95 responses as hallmarks of elevated lysosomal damage.

A further connectivity between the retromer system and membrane atg8ylation was observed at the level of LC3 puncta formation elicited by noncanonical triggers such as LLOMe treatment, which induces CASM and lysosomal repair. The effects of VPS35 on lysosomal functionality and morphology have been previously noted 98,99, and have been ascribed to the effects of the retromer complex on the recycling of the sorting receptors for lysosomal hydrolases, such as cation-independent mannose 6-phosphate receptor (CI-MPR) 98. We have detected significant increase in LC3 puncta in VPS35 KO cells stimulated for autophagy by starvation, consistent with the reported diminished lysosomal degradative capacity when VPS35 is absent, an effect ascribed to aberrant CI-MPR sorting 98,99. Whether CI-MPR is sorted by retromer or independently by the ESCPE-I complex 70 remains controversial 57,70,75,98,110,118–120. This controversy has been extended to the sorting of cation-dependent mannose 6-phosphate receptor (CD-MPR) 98,121, further confounded by the fact that retriever, which participates in receptor recycling, shares VPS29 subunit with the retromer 61.

In this study, we observed a miss-localization of the small GTPase Rab7 to lysosomes in ATG5 KO cells, paralleling the effects on GLUT1. Retromer and SNX adaptors (SNX3) interact with Rab7 108. Rab7 affects retromer-dependent sorting 106,108,109,122. Depletion of Rab7 reduces endosomal association of retromer 108 whereas retromer status affects Rab7, i.e., in cells with retromer subunits knocked out, Rab7 anomalously accumulates on lysosomes 109. Consistent with this, we observed that ATG5 KO phenocopied effects of VPS35 KO and caused Rab7 re-localization to lysosomal membranes. Rab7, like other Rabs 114, is controlled by GEFs (Mon1-Cz1), GAPs (ArmusTBC1D2A, TBC1D2B, TBC1D15), and GDI 115. It is curious that all Rab7 GAPs have LC3 interaction region (LIR) motifs and bind mATG8s: Armus/TBC12DA and TBC1D5 bind LC3A 106,113, GABARAPL1 113, and LC3C 106; TBCD12B binds all mATG8s 113,123; and TBC1D15 binds LC3A 124, LC3B, LC3C, GABARAP, GABARAP L1 and GABARAPL2 123,124. Prior elegant studies have shown that genetic ablation of the Rab7 GAP TBC1D5 results in hyperactivation of Rab7 in a wrong intracellular locale (lysosome) 109. In our work, carried out under basal conditions or in cells subjected to lysosomal damage, we observed both translocation of Rab7 to and entrapment of GLUT1 on lysosomes. However, we did not detect any changes in TBC1D5-mATG8s interactions, suggesting that sequestration of this Rab7 GAP by mATG8s, leading to increased plasma membrane localization of GLUT1 under glucose starvation conditions 106, is not at play here. Furthermore, we found that overexpression of constitutively active Rab7 in cells grown in glucose/nutrient-rich media causes lysosomal retention of GLUT1. This is consistent with the reports by others that perturbances in the retromer-TBC1D5 complex lead to miss-localization of Rab7 to more lysosomal-like subdomains within the endosomal system 110 and that inactivation of Rab7’s GAP TBC1D5 leads to entrapment within the endosomal system of the plasma membrane receptors as well as receptors that normally cycle back to the trans-Golgi network 125.

The role of ATG5 in retromer-dependent sorting of GLUT1 appears to have two components, the first one reflects ATG5 being a part of the E3 ligases for membrane atg8ylation and the second one reflects ATG5’s being in protein complexes with the core retromer subunits VPS26, VPS29 and VPS35. The latter association potentially explains the stronger effects of ATG5’s absence on GLUT1 sorting relative to the inactivation of other membrane atg8ylation genes. Of note, interactions between ATG5 and retromer are independent of the conjugation status of ATG5, as both ATG12— ATG5 conjugates and unconjugated ATG5 were found in protein immunoprecipitates of retromer components. This is consistent with our LC-MS/MS proteomic data indicating a conjugation status-independent increase in the proximity of ATG5 and retromer subunits during lysosomal damage. One consequence of the association between retromer and ATG5, a key component of the membrane atg8ylation apparatus, is that this could spatially direct proper membrane atg8ylation and be in part responsible for the effects of VPS35 KO on lysosomal quality and sensitivity to membrane stress or injury. In our hands, AlphaFold modeling of ATG5 with retromer components did not yield any high-confidence structures, and the best of the low-to-moderate probability structures when subjected to mutational analysis did not result in disruption of ATG5-retromer association observed by co-IPs (data not shown). Identification of the ATG5 partners that bring it to the vicinity of the retromer is yet to be accomplished and remains one of the limitations of the present study.

We used GLUT1 as a well-defined cargo sorted by the retromer in conjunction with SNX27 72,73. SNX27 is a versatile adapter, linking retromer to and controlling endosome-to-plasma membrane recycling of nearly 80 proteins such as signaling receptors, ion channels, amino acid and other nutrient transporters 73. SNX27 was also missorted in ATG5 deficient cells. This suggests that knockouts in ATG5 and in other membrane atg8ylation genes could have complex pleiotropic effects on cellular functions. Given that ATG5 has a particularly strong impact on retromer-dependent sorting, this could in part explain its unique role in the exquisite susceptibility of Atg5 mutant mice to experimental M. tuberculosis infections. Such effects likely contribute to the specific inflammatory action of neutrophils observed in Atg5fl/fl LysM-Cre mice 35,46,53–56. More narrowly, ATG5 KO-dependent GLUT1 miss-sorting may be one of the contributors to increased M. tuberculosis pathogenesis in infection sites. More generally, glucose uptake and metabolism are highly important for normal physiology and in various disease states including immune responses to infection, cancer, neurodegeneration, cardiovascular health, and metabolic disorders. We postulate that deficiencies or polymorphisms in membrane atg8ylation genes could have contributory roles in health and disease via GLUT1 and additional transporter proteins and signaling receptors that are dependent on retromer for their proper positioning and function in the cell.

In conclusion, we find that the membrane atg8ylation and retromer systems are intertwined, and that they affect each other’s biological outputs, one being resilience of lysosomes to basal stress or induced damage and the other being endosomal-plasma membrane protein sorting, both being of fundamental interest and potential therapeutic value.

eLife Assessment

This important work identifies a non-autophagic role for ATG5 in lysosomal repair and the trafficking of the glucose transporter GLUT1 to the cell surface, mediated through the retromer complex. The evidence supporting the conclusions is solid.

Reviewer #1 (Public Review):

Summary:

In this study, Masroor Ahmad Paddar and his/her colleagues explore the noncanonical roles of ATG5 and membrane atg8ylation in regulating retromer assembly and function. They begin by examining the interactomes of ATG5 and expand the scope of these effects to include homeostatic responses to membrane stress and damage.

Strengths:

This study provides novel insights into the noncanonical function of ATG8ylation in endosomal cargo sorting process.

Weaknesses:

The direct mechanism by which ATG8ylation regulates the retromer remains unsolved.

Reviewer #2 (Public Review):

Summary: Padder et al. demonstrate that ATG5 mediates lysosomal repair via the recruitment of the retromer components during LLOMe-induced lysosomal damage and that mAtg8-ylation contributes to retromer-dependent cargo sorting of GLUT1. Although previous studies have suggested that during glucose withdrawal, classical autophagy contributes to retromer-dependent GLUT1 surface trafficking via interactions between LC3A and TBC1D5, the experiments here demonstrate that during basal conditions or lysosomal damage, ATGs that are not involved in mATG8ylation, such as FIP200, are not functionally required for retromer-dependent sorting of GLUT1. Overall, these studies suggest a unique role for ATG5 in the control of retromer function, and that conjugation of ATG8 to single membranes (CASM) is a partial contributor to these phenotypes.

Strengths:

(1) Overall, these studies suggest a unique non-autophagic role for ATG5 in the control of retromer function. They also demonstrate that conjugation of ATG8 to single membranes (CASM) is a partial contributor to these phenotypes. Overall, these data point to a new role for ATG5 and CASM-dependent mATG8ylation in lysosomal membrane repair and trafficking.

(2) Although the studies are overall supportive of the proposed model that the retromer is controlled by CASM-dependent mATG8-ylaytion, it is noteworthy that previous studies of GLUT1 trafficking during glucose withdrawal (Roy et al. Mol Cell, PMID: 28602638) were predominantly conducted in cells lacking ATG5 or ATG7, which would not be able to discriminate between a CASM-dependent vs. canonical autophagy-dependent pathway in the control of GLUT1 sorting. Is the lack of GLUT1 mis-sorting to lysosomes observed in FIP200 and ATG13KO cells also observed during glucose withdrawal? Notably, deficiencies in glycolysis and glucose-dependent growth have been reported in FIP200 deficient fibroblasts (Wei et al. G&D, PMID: 21764854) so there may be differences in regulation dependent on the stress imposed on a cell.

Weaknesses:

(1) Additional controls are needed to clarify the role of CASM in the control of retromer function. Because the manuscript proposes both CASM-dependent and independent pathways in the ATG5 mediated regulation of the retromer, it is important to provide robust evidence that CASM is required for retromer-dependent GLUT1 sorting to the plasma membrane vs. lysosome. The experiments with monsensin in Fig. 7C-E are consistent with but not unequivocally corroborative of a role for CASM. Based on the results shown with ATG16KO in Fig 4A-D, rescue experiments of these 16KO cells with WT vs. C-terminal WD40 mutant versions of ATG16 will specifically assess the requirement for CASM and potentially provide more rigorous support for the conclusions drawn.

(2) Also, the role of TBC1D5 should be further clarified. In Fig S7, are there any changes in the interactions between TBC1D5 and VPS35 in response to LLOMe or other agents utilized to induce CASM? Does TBC1D5 loss-of-function modulate the numbers of GLUT1 and Gal3 puncta observed in ATG5 deficient cells in response to LLOMe?

(3) Finally, the studies here are motivated by experiments in Fig. S1 (as well as other studies from the Deretic and Stallings labs) suggesting unique autophagy-independent functions for ATG5 in myeloid cells and neutrophils in susceptibility to Mycobacterium tuberculosis infection. However, it is curious that no attempt is made to relate the mechanistic data regarding the retromer or GLUT1 receptor mis-sorting back to the infectious models. Do myeloid cells or neutrophils lacking ATG5 have deficiencies in glucose uptake or GLUT1 cell surface levels?

Reviewer #3 (Public Review):

In this manuscript, Padder et al. used APEX2 proximity labeling to find an interaction between ATG5 and the core components of the Retromer complex, VPS26, VPS29, and VPS35. Further studies revealed that ATG5 KO inhibited the trafficking of GLUT1 to the plasma membrane. They also found that other autophagy genes involved in membrane atg8ylation affected GLUT1 sorting. However, knocking out other essential autophagy genes such as ATG13 and FIP200 did not affect GLUT1 sorting. These findings suggest that ATG5 participates in the function of the Retromer in a noncanonical autophagy manner. Overall, the methods and techniques employed by the authors largely support their conclusions. These findings are intriguing and significant, enriching our understanding of the non-autophagic functions of autophagy proteins and the sorting of GLUT1. Nevertheless, there are several issues that the authors need to address to further clarify their conclusions.

(1) The authors confirmed the interaction between Atg5 and the Retromer complex through Co-IP experiments. Is the interaction between Atg5 and the Retromer direct? If it is direct, which Retromer complex protein regulates the interaction with Atg5? Additionally, does ATG5 K130R mutant enhance its interaction with the Retromer?

(2) To more directly elucidate how ATG5 regulates Retromer function by interacting with the Retromer and participates in the trafficking of GLUT1 to the plasma membrane, the authors should identify which region or crucial amino acid residues of ATG5 regulate its interaction with the Retromer. Additionally, they should test whether mutations in ATG5 that disrupt its interaction with the Retromer affect Retromer function (such as participating in the trafficking of GLUT1 to the plasma membrane) and whether they affect Atg8ylation. They also need to assess whether these mutations influence canonical autophagy and lysosomal sensitivity to damage.

Author response:

The following is the authors’ response to the original reviews.

Public reviews:

Reviewer #1 (Public Review):

Summary:

In this study, Masroor Ahmad Paddar and his/her colleagues explore the noncanonical roles of ATG5 and membrane atg8ylation in regulating retromer assembly and function. They begin by examining the interactomes of ATG5 and expand the scope of these effects to include homeostatic responses to membrane stress and damage.

Strengths:

This study provides novel insights into the noncanonical function of ATG8ylation in endosomal cargo sorting process.

Weaknesses:

The direct mechanism by which ATG8ylation regulates the retromer remains unsolved.

We agree with the reviewer. We do however show how at least one aspect of atg8ylation contributes to the proper retromer function, which occurs via lysosomal membrane maintenance and repair. Understanding the more direct effects on retromer will require a separate study. We now emphasize this in the revised manuscript (p. 18) and point out the limitations of the present work (p. 18): “One of the limitations of our study is that beyond effects of membrane atg8ylation on quality of lysosomal membrane and its homeostasis there could be more direct effects of membrane modification with mATG8s that still need to be understood”.

Reviewer #2 (Public Review):

Summary:

Padder et al. demonstrate that ATG5 mediates lysosomal repair via the recruitment of the retromer components during LLOMe-induced lysosomal damage and that mAtg8-ylation contributes to retromer-dependent cargo sorting of GLUT1. Although previous studies have suggested that during glucose withdrawal, classical autophagy contributes to retromer-dependent GLUT1 surface trafficking via interactions between LC3A and TBC1D5, the experiments here demonstrate that during basal conditions or lysosomal damage, ATGs that are not involved in mATG8ylation, such as FIP200, are not functionally required for retromer-dependent sorting of GLUT1. Overall, these studies suggest a unique role for ATG5 in the control of retromer function, and that conjugation of ATG8 to single membranes (CASM) is a partial contributor to these phenotypes.

Strengths:

(1) Overall, these studies suggest a unique non-autophagic role for ATG5 in the control of retromer function. They also demonstrate that conjugation of ATG8 to single membranes (CASM) is a partial contributor to these phenotypes. Overall, these data point to a new role for ATG5 and CASM-dependent mATG8ylation in lysosomal membrane repair and trafficking.

(2) Although the studies are overall supportive of the proposed model that the retromer is controlled by CASM-dependent mATG8-ylaytion, it is noteworthy that previous studies of GLUT1 trafficking during glucose withdrawal (Roy et al. Mol Cell, PMID: 28602638) were predominantly conducted in cells lacking ATG5 or ATG7, which would not be able to discriminate between a CASM-dependent vs. canonical autophagy-dependent pathway in the control of GLUT1 sorting. Is the lack of GLUT1 mis-sorting to lysosomes observed in FIP200 and ATG13KO cells also observed during glucose withdrawal? Notably, deficiencies in glycolysis and glucose-dependent growth have been reported in FIP200 deficient fibroblasts (Wei et al. G&D, PMID: 21764854) so there may be differences in regulation dependent on the stress imposed on a cell.

We thank the reviewer for the overall assessment of the strengths of the study. We have discussed in the manuscript the elegant study by Roy et al., PMID 28602683. To accommodate reviewer’s comment, we have additionally emphasized in the text that our study is focused on basal conditions and conditions that perturb endolysosomal compartments. We agree with the reviewer that under metabolic stress conditions (such as glucose limitation) more complex pathways may be engaged and have acknowledged that in the discussion. We have now included this in the limitations of the study (p. 18): “Another limitation of our study is that we have focused on basal conditions or conditions causing lysosomal damage, whereas metabolic stress including glucose excess or limitation with its multitude of metabolic effects have not been addressed”.

Weaknesses:

(1) Additional controls are needed to clarify the role of CASM in the control of retromer function. Because the manuscript proposes both CASM-dependent and independent pathways in the ATG5 mediated regulation of the retromer, it is important to provide robust evidence that CASM is required for retromer-dependent GLUT1 sorting to the plasma membrane vs. lysosome. The experiments with monensin in Fig. 7C-E are consistent with but not unequivocally corroborative of a role for CASM.

We fully agree with the reviewer. In fact, our data with bafilomycin A1 treatment causing GLUT1 miss-sorting show that it is the perturbance of lysosomes and not CASM per se that leads to mis-sorting of GLUT1 (Fig. 7D,E). Note that it has been shown (PMIDs: 28296541, 25484071 and 37796195) that although bafilomycin A1 deacidifies lysosomes it does not induce but instead inhibits CASM. This is because bafilomycin A1 causes dissociation of V1 and V0 sectors of V-ATPase, unlike other CASM-inducing agents which promote V1 V0 association. Complementing this, our data with ATG2AB DKO and ESCRT VPS37A KO (Fig. 8A-F) indicate that the repair of lysosomes is important to keep the retromer machinery functional (as illustrated in Fig. 8G). This may be one of the effector mechanisms downstream of membrane atg8ylation in general and hence also downstream of CASM. We have revised Fig. 7 title to read “Lysosomal perturbations cause GLUT1 mis-sorting” and have explained these relationships in the text (p. 12-13): “Since bafilomycin A1 does not induce CASM but disturbs luminal pH, we conclude that it is the less acidic luminal pH of the endolysosomal organelles, and not CASM, that is sufficient to interfere with the proper sorting of GLUT1.”

Based on the results shown with ATG16KO in Fig 4A-D, rescue experiments of these 16KO cells with WT vs. C-terminal WD40 mutant versions of ATG16 will specifically assess the requirement for CASM and potentially provide more rigorous support for the conclusions drawn.

We have carried out complementation with ATG16L1 WT and its E230 mutant (devoid of WD40 repeats but still capable of canonical autophagy) and placed these data in Fig. 7 (panels I and J) as recommended by the reviewer. This is now described on p. 13 (To additionally test this notion, we compared ATG16L1 full length (ATG16L1FL) and ATG16L1E230 (Rai et al., PMID 30403914) for complementation of the GLUT1 sorting defect in ATG16L1 KO cells (Fig. 7I,J). ATG16L1E230 [Rai, 2019, 30403914] lacks the key domain to carry out CASM via binding to VATPase 29,30 31-33 but retains capacity to carry out atg8ylation. Both ATG16L1FL and ATG16L1E230 complemented mis-sorting of GLUT1 (Fig. 7I,J). Collectively, these data indicate that it is not absence of CASM/VAIL but absence of membrane atg8ylation in general that promotes GLUT1 mis-sorting.).

(2) Also, the role of TBC1D5 should be further clarified. In Fig S7, are there any changes in the interactions between TBC1D5 and VPS35 in response to LLOMe or other agents utilized to induce CASM?

We thank the reviewer for pointing this out. We do have data with VPS35 in co-IPs shown in Fig. S7. There is no change in the amounts of VPS35 or TBC1D5 in GFP-LC3A co-IPs. We now include in Fig. S7 (new panel D) a graph with quantification in the revised manuscript and emphasize this point (p. 12): “However, under CASM-inducing conditions, no changes were detected (Fig. S7B-D) in interactions between TBC1D5 and LC3A or in levels of VPS35 in LC3A co-IP, a proxy for LC3A-TBC1D5-VPS29/retromer association. This suggests that CASM-inducing treatments and additionally bafilomycin A1 do not affect the status of the TBC1D5-Rab7 system”.

Does TBC1D5 loss-of-function modulate the numbers of GLUT1 and Gal3 puncta observed in ATG5 deficient cells in response to LLOMe?

We agree that TBC1D5 is an interesting aspect. However, because TBC1D5 does not change its interactions in the experiments in our study, we consider this topic (i.e. whether TBC1D5 phenocopies VPS35 and ATG5 KOs in its effects on Gal3) to be beyond the scope of the present work. We underscore that LLOMe (lysosomal damage) mis-sorts GLUT1 even without any genetic intervention (e.g., in WT cells in the absence of ATG5 KO; Fig. 7). Thus, in our opinion the effects of TBC1D5 inactivation may be a moot point.

(3) Finally, the studies here are motivated by experiments in Fig. S1 (as well as other studies from the Deretic and Stallings labs) suggesting unique autophagy-independent functions for ATG5 in myeloid cells and neutrophils in susceptibility to Mycobacterium tuberculosis infection. However, it is curious that no attempt is made to relate the mechanistic data regarding the retromer or GLUT1 receptor mis-sorting back to the infectious models. Do myeloid cells or neutrophils lacking ATG5 have deficiencies in glucose uptake or GLUT1 cell surface levels?

Reviewer’s point is well taken. Glucose uptake, its metabolism, and diabetes underly resurgence in TB in certain populations and are important factors in a range of other diseases. This was alluded to in our discussion (lines 461-469). However, these are complex topics for future studies. We have now expanded this section of the discussion (p. 18): “In the context of tuberculosis, diabetes, which includes glucose dysregulation, is associated with increased incidence of active disease and adverse outcomes” (Dheda et al., ,PMID: 26377143; Dooley, et al., PMID:19926034).

Reviewer #3 (Public Review):

In this manuscript, Padder et al. used APEX2 proximity labeling to find an interaction between ATG5 and the core components of the Retromer complex, VPS26, VPS29, and VPS35. Further studies revealed that ATG5 KO inhibited the trafficking of GLUT1 to the plasma membrane. They also found that other autophagy genes involved in membrane atg8ylation affected GLUT1 sorting. However, knocking out other essential autophagy genes such as ATG13 and FIP200 did not affect GLUT1 sorting. These findings suggest that ATG5 participates in the function of the Retromer in a noncanonical autophagy manner. Overall, the methods and techniques employed by the authors largely support their conclusions. These findings are intriguing and significant, enriching our understanding of the non-autophagic functions of autophagy proteins and the sorting of GLUT1.

Nevertheless, there are several issues that the authors need to address to further clarify their conclusions.

(1) The authors confirmed the interaction between Atg5 and the Retromer complex through Co-IP experiments. Is the interaction between Atg5 and the Retromer direct? If it is direct, which Retromer complex protein regulates the interaction with Atg5? Additionally, does ATG5 K130R mutant enhance its interaction with the Retromer?

AlphaFold modeling in the initial submission of our study to eLife (absent from the current version) suggested the possibility of a direct interaction between ATG5 and VPS35 with ATG12—ATG5 complex facing outwards, in which case K130R would not matter. However, mutational experiments in putative contact residues did not alter association in co-IPs. So either ATG5 interacts with other retromer subunits or more likely is in a larger protein complex containing retromer. It will take a separate study to dissect associations and find direct interaction partners.

(2) To more directly elucidate how ATG5 regulates Retromer function by interacting with the Retromer and participates in the trafficking of GLUT1 to the plasma membrane, the authors should identify which region or crucial amino acid residues of ATG5 regulate its interaction with the Retromer. Additionally, they should test whether mutations in ATG5 that disrupt its interaction with the Retromer affect Retromer function (such as participating in the trafficking of GLUT1 to the plasma membrane) and whether they affect Atg8ylation. They also need to assess whether these mutations influence canonical autophagy and lysosomal sensitivity to damage.

Please see the response to point 1.

Recommendations for the authors.

Reviewer #1 (Recommendations For The Authors):

While most data are solid and convincing, the following questions need to be addressed before publication:

Major Concerns:

(1) Examining only one cargo (GLUT1) is insufficient to reflect the retromer's function comprehensively. At least two additional cargoes should be analyzed to observe the phenotypes more accurately.

We agree that having another retromer cargo (in addition to GLUT1) would be of interest. We point out that our data also show mis-sorting of SNX27 to lysosomes (Fig. 3H, quantifications in Fig. 3I). SNX27 in turn sorts nearly 80 ion channels, signaling receptors, and other nutrient transporters. Which of the 80 cargos to prioritize and check (the expectation is that all 80 might be missorted given that they need SNX27)? We have instead tested MPR, a SNX27-independent cargo. We now include data on effects of ATG5 knockout on CI-MPR (Fig. S9A-F). This is described in the text (p. 14; “Effect of ATG5 knockout on MPR sorting

We tested whether ATG5 affects cation-independent mannose 6-phosphate receptor (CI-MPR). For this, we employed the previously developed methods (Fig. S9A) of monitoring retrograde trafficking of CI-MPR from the plasma membrane to the TGN 70,118-121. In the majority of such studies, CI-MPR antibody is allowed to bind to the extracellular domain of CI-MPR at the plasma membrane and its localization dynamics following endocytosis serves as a proxy for trafficking of CI-MPR. We used ATG5 KOs in HeLa and Huh7 cells and quantified by HCM retrograde trafficking to TGN of antibody-labeled CI-MPR at the cell surface, after being taken up by endocytosis and allowed to undergo intracellular sorting, followed by fixation and staining with TGN46 antibody. There was a minor but statistically significant reduction in CIMPR overlap with TGN46 in HeLaATG5-KO that was comparable to the reduction in HeLa cells when

VPS35 was depleted by CRISPR (HeLaVPS35-KO) (Fig. S9B,C). Morphologically, endocytosed Ab-CI-

MPR appeared dispersed in both HeLaATG5-KO and HeLaVPS35-KO cells relative to HeLaWT cells (Fig. S9D). Similar HCM results were obtained with Huh7 cells (WT vs. ATG5KO; Fig. S9E,F). We interpret these data as evidence of indirect action of ATG5 KO on CI-MPR sorting via membrane homeostasis, although we cannot exclude a direct sorting role via retromer. We favor the former interpretation based on the strength of the effect and the controversial nature of retromer engagement in sorting of CI-MPR (57,70,75,98,120).”)

(2) The evidence from Alphafold predictions is weak. The direct interaction of ATG5 with retromer subunits should be tested.

Please see the above response to Reviewer 3.

In addition, does retromer also interact with ATG16L1 similarly to the phenomenon in VAIL?

We fully agree with the reviewer that finding the direct interacting partners between retromer and membrane atg8ylation machinery is an important direction as in our opinion it would expand the repertoire of E3 ligases and its adaptors. However, given the complexity and variety of possibilities, we believe that this is a topic for a future study.

(3) In Line 166, Figures 2C and 2D, the Gal3 phenotype does not seem to be well complemented by VPS35.

We have adjusted the text to acknowledge incomplete complementation (p.7).

(4) In Figures 3 and 4, the authors show that KO of membrane atg8ylation machineries and ATG8-Hexa KO affects the localization of retromer cargo GLUT1 and SNX27. However, the mechanism by which membrane ATG8ylation affects retromer remains unresolved.

Additionally, are other retromer subunits' locations are also affected, if so, how are they impacted? At least a speculative explanation should be provided.

Following reviewers request, we now state on p. 19 that “one of the limitations of our study is that beyond effects of membrane atg8ylation on quality of lysosomal membrane and its homeostasis there could be more direct effects of membrane modification with mATG8s on retromer that still need to be understood”.

(5) In Figure 3, endogenous IP results are required to examine the interaction of ATG5 with retromer if suitable retromer antibodies for IP are available.

Endogenous IPs are given in Fig. 1. We have modified text on p. 8 to clarify this.

(6) In Figure 4, ATG8 Hexa KO, and triple KO of LC3s or GABARAPs all increase the localization of GLUT1 on lysosomes. It seems redundant for ATG8 family proteins here.

Can any individual member of the ATG8 family rescue this phenotype?

If the intent of such complementation analysis is to identify a specific mATG8 responsible for the observed effects, this is already pre-empted by the fact that TKOs also have a similar effect as HEXA mutants (i.e. loss of at least two of mATG8s is enough to cause the phenotype). We now discuss this in the text (p. 10): “Thus, at least two mATG8s, each one from two different mATG8 subclasses (LC3s and GABARAPs) or the entire membrane atg8ylation machinery was engaged in and required for proper GLUT-1 sorting”.

(7) In Figure 5, knockdown of ATG5 in FIP200 KO cells inhibited GLUT1 sorting from endosomes, leading to its trafficking to lysosomes. However, it is known that very little remnant ATG5 in ATG5 KD cells is enough to support ATG8 lipidation. Therefore, it is essential to repeat this experiment using ATG5/FIP200 double KO or ATG5 KO combined with an autophagy inhibitor.

We point out to this limitation in the text (p. 11): “….we knocked down ATG5 in FIP200 KO cells (Fig. S5D) and found that GLUT1 puncta and GLUT1+LAMP2+ profiles increased even in the FIP200 KO background with the effects nearing those of VPS35 knockout (Figs. 5D-F and S5C), with the difference between VPS35 KO and ATG5 KD attributable to any residual ATG5 levels in cells subjected to siRNA knockdowns”.

(8) In Figure 7, the authors show that the induction of CASM inhibited GLUT1 sorting from endosomes. However, ATG5 KO, which abolishes membrane ATG8ylation, also inhibits GLUT1 sorting. This seems paradoxical and requires a reasonable explanation or discussion.

We understand reviewer’s comment. The answer to this paradox is that it is actually the lysosomal damage that causes GLUT1 mis-sorting and not CASM. Membrane atg8ylation, such as CASM and probably other processes given that involvement of both ATG2 and ESCRTs (Fig. 8) counteracts the damage and works in the direction of restoring/maintaining proper retromer-dependent sorting. This is now explained better in the text, and have revised the title of Fig. 7 to read “Lysosomal damage causes GLUT1 mis-sorting”. Our data with bafilomycin A1 show that it is the perturbance of lysosomes (not CASM per se) that leads to mis-sorting of GLUT1 (Fig. 7D,E), and our data with ATG2AB DKO and ESCRT (VPS37A) KO (Fig. 8A-F) indicate that repair of lysosomes is important to keep the retromer working machinery functional (as illustrated in Fig. 8G), which may be one of the effector mechanisms downstream of membrane atg8ylation in general (and hence also of CASM).

(9) The immuno-staining results for Figures 7F and 7G are lacking.

We now provide the requested images.

(10) In Figure 8D, the quality of the image for VPS37 KO cells treated with LLOME is not sufficient to show increased colocalization between GLUT1 and LAMP2.

We now provide a different example image. We note that these are epiflorescent HCM images

Minor Concerns:

(1) It would be better to distinguish the function of the membrane ATG8ylation machinery (i.e., ATG5) from the function of membrane ATG8ylation in the description. No ATG8ylation-deficient mutants were used in this study.

We have used atg8ylation mutants (e.g. KOs in ATG3, ATG5, ATG7, and ATG16L1). We now emphasize this better in the text (p. 10).

(2) In Figure 2D, a green box appears there by incident.

This has been fixed.

(3) In Figure 3A, the conjugate for ATG5-ATG12 is absent in the gel for IB: ATG5.

The ATG5 antibody used in Fig. 3A recognizes primarily the conjugated form of ATG5. This is now clarified in the figure legend.

(4) Figure 5G is missing in the manuscript.

Fig 5G is now mentioned in the text. Thank you.

(5) The gRNA sequence information for FIP200 KO is missing in the Methods section.

Reference(s) to the already published gRNA sequence are in the manuscript.

(6) Suggest moving the last paragraph in Result section to Discussion section.

We kept this single-paragraph section in Results as it contains actual data.

Reviewer #2 (Recommendations For The Authors):

(1) It is unclear why the rescue of VPS35KO cells in Fig 1C-D is so modest.

Complementation data depend on transfection efficiency and some variability is to be expected.

Reviewer #3 (Recommendations For The Authors):

(1) Figures 2A, 2C, 2E, and 2G lack scale bars. Figure 2D has a small square above the y axis.

Relative scale bars are now included.

(2) Figures S3B, S3D, and S3F lack scale bars.

Relative scale bars are now included.

Version 2

Abstract

Abstract

ATG5 is one of the core autophagy proteins with additional functions such as noncanonical membrane atg8ylation, which among a growing number of biological outputs includes control of tuberculosis in animal models. Here we show that ATG5 associates with retromer’s core components VPS26, VPS29 and VPS35 and modulates retromer function. Knockout of ATG5 blocked trafficking of a key glucose transporter sorted by the retromer, GLUT1, to the plasma membrane. Knockouts of other genes essential for membrane atg8ylation, of which ATG5 is a component, affected GLUT1 sorting, indicating that membrane atg8ylation as a process affects retromer function and endosomal sorting. The contribution of membrane atg8ylation to retromer function in GLUT1 sorting was independent of canonical autophagy. These findings expand the scope of membrane atg8ylation to specific sorting processes in the cell dependent on the retromer and its known interactors.

Introduction

The canonical autophagy pathway, ubiquitous in eukaryotes, is manifested by the emergence of intracellular membranous organelles termed autophagosomes that capture cytoplasmic cargo destined for degradation in lysosomes 1. Progress has been made in understanding molecular mechanisms governing canonical autophagosome biogenesis in mammalian cells 2 including ATG9A-vesicles 3–7 and membranous prophagophores 8 as precursors to LC3-positive phagophores, their reshaping as cups 9 and dynamic interactions with the omegasomes 10,11, phagophore expansion through lipid transfer 12–14, cargo recognition and sequestration 15,16, closure of the phagophore to form a double membrane phagosomes 17–20, and fusion of autophagosomes with endosomal and lysosomal organelles leading to cargo degradation 21 or secretion 22.

However, and contemporaneously with these major advances in understanding canonical autophagy, it has become evident that autophagy genes (ATGs) 23 have additional functions in mammalian cells that do not fit the canonical model 24,25. Among such noncanonical manifestations are LAP (LC3-associated phagocytosis) 26, LANDO (LC3-associated endocytosis) 27, LC3-associated micropinocytosis (LAM) 28, conjugation of ATG8 to single membranes (CASM) 29,30, v-ATPase-ATG16L1-induced LC3 lipidation (VAIL) 31–33, ER-phagy mediated by the V-ATPase-ATG16L1-LC3C axis (EVAC) 34, LyHYP (lysosomal hypersensitivity phenotype) 35, membrane damage repair 36–39, and two distinct forms of secretory autophagy 22, SALI (secretory autophagy during lysosome inhibition) 40 and LDELS (LC3-dependent EV loading and secretion) 41. These processes depend on or are associated with the phospholipid conjugation cascade of mammalian ATG8 proteins (mATG8s) 42 and collectively (including the canonical autophagy) represent diverse manifestations of membrane atg8ylation as a broad membrane stress, damage and remodeling response 25.

The factors governing membrane atg8ylation include two enzymatic cascades with the ATG12-ATG5 and mATG8-phosphatidylethanolamine (PE) covalent conjugates as their products. The ATG12-ATG5 conjugate 42 combines with ATG16L1 43 or TECPR1 to form E3 ligases 38,39,44 to direct mATG8-PE conjugation and atg8ylation of target membranes. All known E3 enzymes contain the ATG12-ATG5 conjugate 38,39,42,44. To form this conjugate, the ubiquitin like molecule ATG12 is activated by ATP, and transferred via E1 (ATG7) and E2 (ATG10) to ATG5. In preparation for the next step, an ATG12-ATG5 containing E3 ligase activates its substrate mATG8-ATG3 to transfer ATG8 and form an amide bond with aminophospholipids 42. However, there are additional branches of these conjugation cascades, whereby ATG12 can make a noncanonical sidestep conjugate with ATG3 (ATG12-ATG3) 41,45 which is enhanced in the absence of ATG5 35. Apart from its role in atg8ylation, ATG5 35,46 and potentially other ATG genes 47–49 have autophagy-independent functions. Many of the indications that such functions exist come from in vivo studies 50–52. For example, in the case of murine models of Mycobacterium tuberculosis infection, atg8ylation machinery protects against tuberculosis pathogenesis but Atg5 knockout has a particularly strong phenotype exceeding other atg8ylation genes, sugestiong that ATG5 possesses atg8ylation (mATG8s lipid conjugation) independent functions 35,46,53–56. As recently reported, one such function involves LyHYP contributing to enhanced exocytosis specifically in cells devoid of ATG5 but not of other ATG genes 35. These developments indicate that ATG5 plays noncanonical roles in autophagy-independent vesicular transport events and homeostasis of the endolysosomal system.

Within the endosomal network, the mammalian retromer complex 57 is one of the several systems regulating vectoral transport of membranes and proteins. These systems include, among others, HOPS, CORVET, retriever, CCC/commander, and retromer 57–64 complexes that often contain and sometimes share VPS subunits. Retromer is a heterotrimer of VPS26A/B, VPS29 and VPS35 subunits 65, first identified in yeast 66. Retromer sorts an array of endosomal cargo in cooperation with cognate sorting nexins (SNX) 67. Specifically, this includes SNX-BAR proteins SNX1/2-SNX5/6 68,69, the mammalian paralogs of yeast Vps5p and Vps17p 66 which can function autonomously as an ESCPE-I complex 70, SNX-PX protein SNX3 71, and the SNX-FERM protein SNX27 containing PX, PDZ and FERM domains 72 which sorts various transporter proteins and signaling receptors including the glucose transporter GLUT1 (SLC2A1) 73. Retromer, together with adaptors, contributes to the complex protein sorting within the endosomal system 74–76. Here, using unbiased proteomic approaches and follow-up mechanistic analyses, we report that retromer complex is among ATG5 interactors. We show that membrane atg8ylation, of which ATG5 is an essential component, affects retromer-dependent cargo sorting. This function is independent of the canonical autophagy pathway or a well-studied form of noncanonical membrane atg8ylation, CASM. Our findings offer a paradigm shift connecting ATG5 and membrane atg8ylation with the retromer system and its function in the endosomal cargo sorting, expanding the scope of atg8ylation machinery and its special component ATG5 beyond the current models.

Results

ATG5 associates with the retromer complex

The unique aspects of Atg5 (Fig. 1A), outside of its role in canonical autophagy, have strong ex vivo and important in vivo phenotypes 52. Inactivation of Atg5 in myeloid lineage renders mice excessively susceptible to acute infection with Mycobacterium tuberculosis attributed to excessive inflammation 35,46,53–56, which did not extend to other phases of infection as tested here in a murine model of latent tuberculosis 77–79 (Fig. S1A-D). In a previous study, we could only partially explain the cellular parameters associated with acute infection, which included degranulation in neutrophils from Atg5fl/fl LysM-Cre+ mice 35. Furthermore, markers of LyHYP (lysosome hypersensitivity phenotype) 35 in ATG5 knockout (KO) cells, such as galectin 3 (Gal3) response, could not be explained by the exocytic phenomena 35. We thus sought to uncover additional processes (Fig. 1A) affected by ATG5 at the intracellular level. To include factors beyond the known processes, we analyzed proteomic data obtained with APEX2-ATG5 35 and compared APEX2-ATG5WT with APEX2-ATG5K130R (an ATG5 mutant deficient in conjugation to ATG12) in cells treated with LLOMe, an agent routinely used to cause lysosomal damage 80–85 (Figs. 1B and S2A,B). The proximity biotinylation (APEX2)-based proteomic data indicated the presence in the vicinity of APEX2-ATG5 of several protein complexes that control important membrane trafficking pathways including retromer (Fig. 1B). Beside the retromer VPS subunits, there were very few other VPS proteins (Fig. S2C). VPS proteins were observed in ATG5 proteomic data by others 86 (Fig. S2D). VPS26 and VPS29 showed statistically significant increase in proximity to ATG5 in cells subjected to LLOMe treatment (Fig. S2A-C). Using APEX2-SGALS1 proximity biotinylation LC-MS/MS as a control, we did not observe enrichment of retromer subunits in cells treated with LLOMe (Fig. S2E).

ATG5 interacts with retromer.

A. ATG5 functions. X, a postulated additional function. B. 2D scatter plot (log2 fold changes; color coded p value cutoff matrix for comparisons between samples as per the lookup table) of proximity biotinylation LC/MS/MS datasets: FlpIn-HeLaAPEX2-ATG5-WT cells (X-axis) and FlpIn-HeLaAPEX2-ATG5-K130R (Y-axis) treated with 2 mM LLOMe for 30 min ratioed vs. HeLaAPEX2-ATG5-WT without LLOMe treatment (Ctrl). C-E. Co-IP analyses and quantification of VPS26A (C), VPS29 (D) VPS35, (E), interaction with ATG5 in HeLa cells treated with or without 2 mM LLOMe for 30 min. Data, means ± SE (n=3); unpaired t-test; p values indicated above the bars.

Association of endogenous ATG5 with endogenous retromer subunits (VPS26A, VPS29 and VPS35) was tested in coimmunoprecipitation (co-IP) experiments (Fig. 1C-E). The interaction between retromer and ATG5 was detected in these experiments and showed increased association in cells subjected to lysosomal damage with LLOMe (Fig. 1C-E). Thus, ATG5, a key component of the known atg8ylation E3 ligases 42,87, is found in protein complexes with the retromer and this association is enhanced upon lysosomal damage.

Retromer affects a subset of responses to lysosomal damage

During lysosomal damage a number of processes are set in motion to repair, remove and regenerate/replenish lysosomes 37,80,88–91 including membrane atg8ylation-dependent processes of repair by ESCRT machinery 92 and by lipid transfer via ATG2 85,93. We have reported that ATG5 KO cells display elevated Gal3 response to lysosomal damage agents including LLOMe 35. Gal3 is one of the well-characterized sentinel galectins alerting cellular homeostatic systems to lysosomal damage 37,80,81,88,94. Based on the observed interactions between ATG5 and retromer, we tested whether retromer, like ATG5 35, affected Gal3 response to lysosomal damage. We used the previously established quantitative high content microscopy (HCM) approach, which provides unbiased machine-driven image acquisition and data analysis 8,36,80,88,94. In the experiments herein HCM was based on >500 valid primary objects/cells per well and a minimum of 5 wells per sample, with the independent biological replicates being n≥3 (≥3 separate plates) as described 35. Upon treatment with LLOMe, HeLaVPS35-KO cells had increased Gal3 puncta relative to HeLaWT cells, comparable to HeLaATG5-KO (Fig. 2A,B) 35. The Gal3 phenotype in VPS35 knockout cells was partially complemented by expression of GFP-VPS35 (Fig. 2C,D). The observed elevated Gal3 recruitment to damaged lysosomes in the absence of VPS35 cannot be explained by increased pools of LAMP2 organelles in the cells, since the overall complement of lysosomes (quantified by LAMP2 puncta/cell) did not increase in LLOMe treated HeLa cells (Fig. S3A,B). Another marker of lysosomal damage, ubiquitination response 95, was elevated in both VPS35 and ATG5 deficient cells (Figs. 2E,F). This phenotype was confirmed in another cell line, Huh7 (Fig. S3C,D). Thus, VPS35 and ATG5 defects have a similar effect on lysosomes, previously characterized and termed lysosomal hypersensitivity phenotype/LyHYP 35.

Retromer affects lysosomal sensitivity to damage.

A,B. High content microscopy (HCM) imaging and quantification of Gal3 response (puncta/cell of endogenous Gal3 profiles stained for immunofluorescence) in HeLaWT, HeLaATG5-KO, and HeLaVPS35-KO cells subjected to lysosomal damage by Leu-Leu-O-Me ester hydrobromide (LLOMe; 2mM, 30 min). HCM, an unbiased machine-driven image acquisition and data analysis based on presets of >500 valid primary objects/cells per well (representative images shown; white mask, cell; red masks, Gal3 puncta), with a minimum of 5 wells per sample (sampling error), and n≥3, independent biological replicates (experimental error) in separate 96-well plates. Data, means ± SE (n=3), two-way ANOVA with Tukey’s multiple comparisons. C,D. Complementation analysis of VPS35KO LyHYP (lysosome hypersensitivity) phenotype monitored by HCM quantification of Gal3 puncta in HeLaVPS35-KO cells transfected with GFP (control) or GFP-VPS35 expressing plasmids. Data, means ± SE (n=3); one-way ANOVA with Tukey’s multiple comparisons. E,F. Comparative HCM analysis of ubiquitin (immunofluorescence; FK2 antibody) response to lysosomal damage (LLOMe; 2mM, 30 min) in HeLaATG5-KO and HeLaVPS35-KO cells. Yellow profiles, colocalization of ubiquitin and LAMP1. Top panels, ubiquitin immunostaining alone. Data, means ± SE (n=4), two-way ANOVA with Tukey’s multiple comparisons. G,H. HCM quantification of ALIX localization to endolysosomal compartments (% of LAMP1 profiles positive for ALIX immunostaining) in Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells following lysosomal damage. Yellow profiles, colocalization of ALIX and LAMP1. Data, means ± SE (n=3); two-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line/condition.

Not all aspects of LyHYP, previously observed in ATG5 knockouts 35, were seen in VPS35 KO cells. In the case of ATG5 inactivation (Huh7ATG5-KO and HeLaATG5-KO), the recruitment of ALIX, an ESCRT component involved in early stages of lysosomal repair 80,96,97 is abrogated 35, whereas in VPS35 KO cells (Huh7VPS35-KO and HeLaVPS35-KO) ALIX was efficiently recruited to lysosomes upon LLOMe treatment (Figs. 2G,H and S3E,F).

The elevated Gal3 and ubiquitin markers of increased lysosomal damage in cells with inactivated VPS35 or ATG5 suggest that both may play a related role in lysosomal resilience to membrane damage, which complements the previously observed effects of aberrant VPS35/retromer on lysosomal morphology and proteolytic capacity 98,99. The dissimilarities between ATG5 and retromer effects on the ALIX recruitment component can be explained by the previously described perturbed function of the ATG conjugation machinery specifically in ATG5 KO cells affirming the unique additional roles of ATG5 35. Nevertheless, the striking similarities between ATG5 KO and VPS35 KO effects on heightened Gal3 and ubiquitin responses to lysosomal membrane damage suggest that the two systems (membrane atg8ylation and retromer) closely intersect.

ATG5 affects retromer function

To avoid the confounding, and possibly indirect, effects of lysosomal damage on endolysosomal sorting and trafficking in cells treated with LLOMe, it was necessary to test the effects of ATG5 KO on retromer function under basal conditions with unperturbed lysosomes. Our co-IP analyses with endogenous proteins (Fig. 1C-E) indicated that ATG5 can be found in protein complexes with retromer subunits even without the lysosomal damage. We further confirmed this in co-IPs showing that GFP-VPS35 and YFP-VPS29 were found in complexes with endogenous ATG5 in resting cells, whereas an isotype IgG control did not pulldown retromer components (Fig. 3A). Reverse co-IPs with GFP-VPS35 and YFP-VPS29 detected endogenous ATG5 and ATG12-ATG5 complexes in GFP/YFP immunoprecipitates (Fig. 3B).

ATG5 affects sorting of the retromer cargo GLUT1.

A. Co-IP analysis of endogenous ATG5 with GFP-VPS35 or YFP-VPS29 (transient transfection). B. Reverse Co-IP analysis GFP-VPS35 or YFP-VPS29 (transient transfection) and endogenous ATG5. Note that different antibodies were used in panels A and B: ab108327 (Abcam) recognizing preferentially the conjugated form of ATG5 and abASA-B0113 (Novateinbio), recognizing both conjugated and unconjugated ATG5. C. Confocal images illustrating localization of GLUT1 and LAMP2 in Huh7WT, Huh7VPS35-KO, and Huh7ATG5-KO cells. Scale bar, 10 μm. D,E. HCM quantification of GLUT1 (endogenous protein immunostaining) puncta/cell (D) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (E) in Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. F,G. HCM quantification of GLUT1-SNX27 overlap (% of SNX7 area positive for GLUT1) in Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. H,I. Immunoblot analysis and quantification of proteins in lysosomes purified/enriched by LysoIP (immunoisolation with TMEM192-3xHA) from Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. TMEM192-2xFLAG, negative control. Data, means ± SE (n=3), one-way ANOVA with Tukey’s multiple comparisons. HCM images in panel F, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

Having established that ATG5 and retromer interact under basal conditions, we tested the effects of ATG5 KO on localization of a well-established cargo for retromer-dependent sorting, GLUT1 73,100. We used Huh7 cells, where we generated a panel of membrane atg8ylation and VPS35 CRISPR knockouts (Fig. S4A), because they have a well-defined localization of GLUT1 at the plasma membrane or as intracellular profiles (Fig. 3C), fully amenable (and comparatively more reliable than in HeLa cells) to unbiased quantification by HCM. ATG5 KO in Huh7 cells resulted in redistribution of GLUT1 from the plasma membrane to intracellular punctate profiles (Figs. 3D and S4B). This phenotype was identical to the one observed in VPS35 KO (Fig. 3D and S4B). The increased intracellular GLUT1 puncta overlapped with the lysosomal marker LAMP2 in both ATG5 KO and VPS35 KO cells (Figs. 3C,E and S4C). A similar albeit morphologically less distinct phenotype was observed in HeLa cells (Figs. S4D-H). GLUT1 is sorted by retromer via VPS26’s interactor SNX27, with GLUT1 (and other similar cargo) being captured by the PDZ domain of SNX27 72,73. In WT cells, SNX27 was found in abundant intracellular profiles (Fig. 3F). In both ATG5 KO and VPS35 KO cells, GLUT1 and SNX27 relocalized to the same intracellular compartment (Fig. 3F,G; S4I,J). We carried out lysosomal purification using the well-established LysoIP method 80,91,94,101, and quantified levels of GLUT1 and SNX27 in lysosomal preparations that were positive for LAMP2 and devoid of GM130 (Golgi) and PDI (ER) (Fig. 3H, I). Both GLUT1 and SNX27 were enriched in lysosomal preparations from ATG5 KO and VPS35 KO cells relative to the WT cells (Fig. 3H, I). These experiments biochemically identify the lysosomes as a compartment to which GLUT1 and SNX27 relocalize in ATG5 KO and VSP35 KO cells. The above findings show that a loss of ATG5 affects retromer-controlled trafficking events.

Membrane atg8ylation apparatus contributes to retromer function in GLUT1 sorting

ATG5 is a part of the membrane atg8ylation machinery 25,42,87, also known under the term ‘LC3 lipidation’ 102, which functions within the canonical autophagy pathway but is also engaged in a wide array of noncanonical processes 25,38,39,44,103,104. We tested whether other known components of the atg8ylation apparatus affect retromer. The KOs in Huh7 cells of the components of the prototypical atg8ylation E3 ligase 25,42 included: ATG5 and ATG16L1, as well as the previously characterized Huh7 lines knocked out for E1 and E2 enzymes, ATG7 35 and ATG3 37. These KOs all caused entrapment of GLUT1 in intracellular punctate profiles (Fig. 4A, B) as well as an increase in GLUT1 colocalization with lysosomes (GLUT1+ LAMP2+ profiles; Fig. 4C,D). We tested whether retromer component levels were altered in atg8ylation mutants and found no change in total cellular levels of VPS26, VPS29 and VPS35, tested in ATG3, ATG5 and ATG7 KO Huh7 cells (Fig. S5A,B) as well as in ATG5 KO HeLa cells (Fig. S5C,D). Thus, the entire ATG16L1-dependent atg8ylation apparatus is required to maintain proper retromer-dependent sorting.

Membrane atg8ylation apparatus affects sorting of the retromer cargo GLUT1.

A,B. HCM quantification of GLUT1 (endogenous protein immunostaining) puncta/cell in Huh7WT, Huh7ATG3-KO, Huh7ATG5-KO, Huh7ATG7-KO, Huh7ATG16-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. C,D. HCM quantification of GLUT1 (endogenous protein immunostaining) colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) in Huh7WT, Huh7ATG3-KO, Huh7ATG5-KO, Huh7ATG7-KO, Huh7ATG16-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. E,F. HCM quantification of GLUT1 (endogenous protein immunostaining) colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) in HeLaWT, and HeLaHexa-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

Furthermore, comparison of the previously characterized HeLaHexa-KO cells 105, with all principal mATG8s inactivated, with their isogenic parental HeLaWT cells, showed increased GLUT1 on lysosomes in HeLaHexa-KO (GLUT1+ LAMP2+ profiles; Fig. 4E,F). We observed a similar phenotype in separate triple knockouts (TKOs) of LC3 subfamily and GABARAP subfamily of mATG8s 105 compared to HeLaHexa-KO (Fig. S5E). Thus, at least two or more mATG8s from two different mATG8 subclasses (LC3s and GABARAPs) and the whole membrane atg8ylation machinery were engaged in and required for proper GLUT-1 sorting.

Canonical autophagy cannot explain effects of membrane atg8ylation on GLUT1 sorting

Membrane atg8ylation and mATGs play roles in diverse processes 25 including canonical autophagy 1. We tested whether the participation of atg8ylation in canonical autophagy, previously reported to affect retromer-dependent cargo sorting under nutrient-limiting conditions 106,107, is the reason for reduced trafficking of GLUT1 in our experiments. We used KOs in Huh7 cells of ATG13 and FIP200/RB1CC1, two obligatory components of the canonical autophagy initiation machinery 1, to test whether canonical autophagy under the basal conditions used in our study was responsible for the effects on GLUT1 sorting. We found that in the Huh7 cells knocked out for ATG13 35 or FIP200 37 there was no increase in intracellular GLUT1 puncta in contrast to the increase observed with the isogenic ATG5 and VPS35 mutants (Fig. 5A,B). This was mirrored by no increase of GLUT1+LAMP2+ profiles in ATG13 and FIP200 KO cells whereas colocalization between GLUT1 and LAMP2A increased in ATG5 and VPS35 mutants (Fig. 5C,D). To further confirm that autophagy initiation mutant cells retained the ability to perturb GLUT1 trafficking due to defective atg8ylation, we knocked down ATG5 in FIP200 KO cells (Fig. S5F) and found that GLUT1 puncta and GLUT1+LAMP2+ profiles increased even in the FIP200 KO background with the effects nearing those of VPS35 knockout (Figs. 5E-G and S5G), with the difference between VPS35 KO and ATG5 KD being attributable to residual ATG5 levels in cells subjected to siRNA knockdowns. Thus, we conclude that atg8ylation but not canonical autophagy affects retromer function under basal conditions and that a functional atg8ylation apparatus is required for proper sorting of the retromer cargo GLUT1.

Canonical autophagy does not affect sorting of the retromer cargo GLUT1.

A,B. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell in in Huh7WT, Huh7ATG13-KO, Huh7FIP200–KO, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. C,D. HCM quantification of GLUT1 (immunostaining of endogenous protein) colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) in in Huh7WT, Huh7ATG13-KO, Huh7FIP200–KO, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. E-G. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell (E) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (F) in Huh7WT, Huh7FIP200–KO, Huh7FIP200–KO + siATG5, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

ATG5 and membrane atg8ylation machinery affect Rab7 localization

Rab7 is considered to be an interactor of the retromer 108. Rab7 translocates in cells lacking the retromer component VPS35 to lysosomes 109 including endolysosomal domains with lysosomaly positioned mTORC1 regulatory machinery 110. We thus tested whether absence of ATG5 and components of the atg8ylation apparatus affected Rab7 localization. ATG5, VPS35 as well as knockouts of all major components of the atg8ylation in Huh7 cells, displayed increased Rab7+ cytoplasmic profiles (Fig. 6A,B) and overlap between Rab7 and lysosomes (LAMP2) (Figs. 6C and S6A). This was confirmed biochemically using lysosomal purification by LysoIP and immunoblotting for Rab7: there was an increase of Rab7 in lysosomal preparations from ATG5 and VPS35 KO cells (Fig. 6D,E).

Membrane atg8ylation machinery is required for proper RAB7 localization.

A-C HCM quantification of Rab7 (endogenous protein immunostaining) puncta/cell (B) and Rab7 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (C) in Huh7WT, Huh7ATG3-KO, Huh7ATG5-KO, Huh7ATG7-KO, Huh7ATG16-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. D-E. Immunoblot analysis and quantification of proteins in lysosomes purified/enriched by LysoIP (immunoisolation with TMEM192-3xHA) from Huh7WT, Huh7ATG5-KO, and Huh7VPS35-KO cells. TMEM192-2xFLAG, negative control. Data, means ± SE (n=3), one-way ANOVA with Tukey’s multiple comparisons. F-H. HCM quantification of Rab7 (endogenous protein immunostaining) puncta/cell (G) and Rab7 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area) (H) in Huh7WT, Huh7ATG13-KO, Huh7FIP200–KO, Huh7ATG5-KO, and Huh7VPS35-KO cells. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line.

This effect extended to other components of the atg8ylation machinery, as KOs in ATG3, ATG7 and ATG16L1 caused a similar effect (Figs. 6A-C and S6A). In contrast, KOs in genes specific for canonical autophagy (ATG13 and FIP200) did not alter intracellular distribution of Rab7 (Figs. 6F-H and S6B). Rab7 distribution was affected by ATG5 knockdown (Fig. S5F) in FIP200 KO cells (Figs. S6C-F). These experiments mirror the effects of ATG5 and atg8ylation on GLUT1 trafficking, showing that ATG5 and atg8ylation machinery but not canonical autophagy is required for proper localization of Rab7.

Agonists of endolysosomal atg8ylation process CASM affect GLUT1 sorting

Membrane atg8ylation has multiple presentations that include not only canonical autophagy but also encompass processes such as CASM 103. We thus asked whether induction of CASM 93 could affect GLUT1 trafficking. Using LLOMe as one of the inducers of CASM 93, we detect increased membrane atg8ylation (LC3 puncta formation) in response to lysosomal damage (Fig. 7A,B). When cells were treated with LLOMe 93 and another CASM inducer, monensin 29,111,112, GLUT1 trafficking to the plasma membrane was negatively affected and instead GLUT1 accumulated intracellularly and trafficked to lysosomes (Figs. 7C-E and S7A). Thus, lysosomal perturbation and induction of CASM may affect retromer-dependent sorting of GLUT1.

Lysosomal perturbations cause GLUT1 mis-sorting.

A,B. HCM imaging and quantification of LC3 response (puncta/cell of endogenous LC3 immunofluorescent profiles) in HeLaWT, and HeLaVPS35-KO cells in response to lysosomal damage by LLOMe (1mM, 30 and 60 min). Scale bar, 20 μm. Data, means ± SE (n=5), one-way ANOVA with Tukey’s multiple comparisons. C-E. HCM quantification of GLUT1 response (puncta/cell of endogenous GLUT1, D) and its localization to endolysosomal compartments (% of LAMP1 profiles positive for GLUT1 immunostaining, E) in Huh7WT treated with or without Monensin (100µM), LLOMe (100µM), and Bafilomycin A1 (100nM) for 45 minutes. Scale bar, 20 μm. Data, means ± SE (n=5), one-way ANOVA with Tukey’s multiple comparisons. F-H. Analysis of GLUT1 puncta/cell (F,G) and its localization to endolysosomal compartments (% of LAMP1 profiles positive for GLUT1 immunostaining, (H) phenotype monitored by HCM quantification in Huh7WT cells transfected with GFP (control) or GFP-Rab7WT, GFP-Rab7Q67L, and GFP-Rab7T22N, expressing plasmids. Scale bar, 20 μm. Data, means ± SE (n=3); one-way ANOVA with Tukey’s multiple comparisons. I-J. Analysis of GLUT1 puncta/cell (I,J) by HCM quantification in Huh7 WT and ATG16L1-KO cells complemented with Flag (control), Flag-ATG16L1FL, or Flag-ATG16L1E230, expressing plasmids. Scale bar, 20 μm. Data, means ± SE (n=5); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line/condition.

It was previously reported that under glucose starvation conditions, LC3A recruits TBC1D5, a GAP for Rab7 113–115, away from this small GTPase 106 and a similar phenomenon could be occurring during CASM. However, under CASM-inducing conditions, no changes were detected (Fig. S7B-D) in interactions between TBC1D5 and LC3A or in levels of VPS35 in LC3A co-IPs, a proxy for LC3A-TBC1D5-VPS29/retromer association. This suggests that CASM-inducing treatments and additionally bafilomycin A1 do not affect the status of the TBC1D5-Rab7 system. Moreover, expression of constitutively active Rab7Q67L did not promote trafficking of GLUT1 to plasma membrane but rather caused accumulation of intracellular GLUT1 puncta (Fig. 7F-H).

Membrane atg8ylation and endolysosomal homeostasis affect GLUT1 sorting

A complication of interpreting CASM as a membrane atg8ylation process responsible for effects on retromer-dependent sorting is that CASM is elicited by lysosomal perturbations using lysosomal pH modifying agents 29,103,111,112, and lysosomal stress could be an independent factor preventing proper sorting of GLUT1. Endolysosomal membrane damage (LLOMe) or perturbations of luminal pH (monensin, bafilomycin A1) negatively affected GLUT1 sorting and caused it to accumulate in lysosomes (Fig. 7C-E). Since bafilomycin A1 does not induce CASM 93 but disturbs luminal pH, we conclude that it is the less acidic luminal pH of the endolysosomal organelles, and not CASM, that is responsible for interference with the proper sorting of GLUT1. To additionally test this, we compared ATG16L1 full length (ATG16L1FL) and ATG16L1E230 116 for complementation of the GLUT1 sorting defect in ATG16L1 KO cells (Fig. 7I,J). ATG16L1E230 116 lacks the key domain to carry out CASM via binding to V-ATPase 29,30 31–33 but retains capacity to carry out atg8ylation. Both ATG16L1FL and ATG16L1E230 complemented mis-sorting of GLUT1 (Fig. 7I,J). Collectively, these data indicate that it is not specifically an absence of CASM/VAIL but the absence of membrane atg8ylation in general that promotes GLUT1 mis-sorting.

The above experiments suggest the role of acidification and integrity of endolysosomal compartments in GLUT1 sorting. Membrane atg8ylation is important for lysosomal repair and homeostasis with specific downstream mechanisms. These processes include: (i) ATG2A, which transfers lipids during lysosomal repair 85 and is engaged on damaged lysosomes upon membrane atg8ylation by LC3A 93, and (ii) ESCRT-based lysosomal membrane repair 80,96,97,117, with membrane atg8ylation specifically recruiting an ESCRT regulator ALG2 92 and ESCRT-I component VPS37A 20. Hence, we tested whether these established systems, involved in maintenance of lysosomal membrane integrity, represented by ATG2 and VPS37, can explain the mechanism underlying the observed effects of membrane atg8ylation on retromer-dependent trafficking. When U2OS ATG2A/ATG2B double KO cells were compared to parental U2OS WT cells, we observed aberrant GLUT1 trafficking, presented as accumulation of cytoplasmic GLUT1 puncta and increased colocalization of GLUT1 with lysosomes (LAMP2A) (Figs. 8A-C and S8A). A similar effect was observed when the previously characterized VPS37A knockout cells 20 were tested (Figs. 8D-F and S8B). These effects were detected with or without induced lysosomal damage (Fig. 8A-F).

Membrane atg8ylation and endolysosomal homeostasis affect retromer-dependent sorting.

A-C. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell (B) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area, C) in U2OSWT and U2OSATG2A/B-DKO cells treated with or without LLOMe (100µM) for 45 minutes. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. D-F. HCM quantification of GLUT1 (immunostaining of endogenous protein) puncta/cell (E) and GLUT1 colocalization with LAMP2 (% of LAMP2 profiles positive for GLUT1; overlap area, F) in Huh7WT cand Huh7VPS37A-KO cells treated with or without LLOMe (100µM) for 45 minutes. Scale bar, 20 μm. Data, means ± SE (n =6); one-way ANOVA with Tukey’s multiple comparisons. HCM images in all relevant panels, examples from a bank of unbiased operator-independent machine-collected and algorithm-processed fields containing a minimum of 500 primary objects/cells per well (5 wells minimum per 96-well plate; 3 plates minimum), per cell line/condition. G. Schematic: Membrane atg8ylation maintains membrane homeostasis under basal or stress conditions and, independently of canonical autophagy, affects retromer function under. A functional atg8ylation apparatus is required for proper sorting of the retromer cargo GLUT1.

We interpret these data as evidence that endolysosomal homeostasis and functionality, which is at all times maintained by membrane atg8ylation, even under basal conditions, contributes to proper function of retromer-dependent sorting of its cognate cargo such as GLUT1 (Fig. 8G).

Effect of ATG5 knockout on MPR sorting

We tested whether ATG5 affects cation-independent mannose 6-phosphate receptor (CI-MPR). For this, we employed the previously developed methods (Fig. S9A) of monitoring retrograde trafficking of CI-MPR from the plasma membrane to the TGN 70,118–121. In the majority of such studies, CI-MPR antibody is allowed to bind to the extracellular domain of CI-MPR at the plasma membrane and its localization dynamics following endocytosis serves as a proxy for trafficking of CI-MPR. We used ATG5 KOs in HeLa and Huh7 cells and quantified by HCM retrograde trafficking to TGN of antibody-labeled CI-MPR at the cell surface, after being taken up by endocytosis and allowed to undergo intracellular sorting, followed by fixation and staining with TGN46 antibody. There was a minor but statistically significant reduction in CI-MPR overlap with TGN46 in HeLaATG5-KO that was comparable to the reduction in HeLa cells when VPS35 was depleted by CRISPR (HeLaVPS35-KO) (Fig. S9B,C). Morphologically, endocytosed Ab-CI-MPR appeared dispersed in both HeLaATG5-KO and HeLaVPS35-KO cells relative to HeLaWT cells (Fig. S9D). Similar HCM results were obtained with Huh7 cells (WT vs. ATG5KO; Fig. S9E,F). We interpret these data as evidence of indirect action of ATG5 KO on CI-MPR sorting via membrane homeostasis, although we cannot exclude a direct sorting role via retromer. We favor the former interpretation based on the magnitude of the effect and the controversial nature of retromer engagement in sorting of CI-MPR 57,70,75,98,120.

ATG5 effects on retromer function exceed those of membrane atg8ylation

Finally, we addressed the strength of effects on GLUT1 trafficking observed with ATG5 KO vs. KOs in other membrane atg8ylation genes, which were significant in all experiments but presented effects lower in magnitude than ATG5 KO (Figs. 4 and 6). One possibility was that since ATG3, ATG5, and ATG7 all participate in the same process of membrane atg8ylation, that our KO mutants had unequal gene inactivation levels. This was ruled out by comparing effects of ATG3 KO, ATG5 KO and ATG7 KO cells for their effects on LC3 puncta formation in a well-defined system of starvation-induced autophagy. The results of these analyses indicated that all three KOs had indistinguishable effects on LC3 puncta formation during canonical autophagy induction by starvation (Fig. S10A,B). Thus, the stronger effects of ATG5 KO on retromer vis-à-vis KOs in other atg8ylation genes reflected additional action of ATG5. In the absence of ATG5, less VPS26 and VPS35 could be pulled down in co-IPs with YFP-VPS29 (Fig. S10C-F). We conclude that ATG5, in addition to contributing to the retromer-dependent sorting via membrane atg8ylation and its downstream effector mechanisms which maintain healthy endolysosomal organelles, has additional effects on the retromer.

Discussion

In this study, we have uncovered a new role of ATG5 and membrane atg8ylation machinery beyond the conventional function in canonical autophagy and the hitherto recognized noncanonical processes. ATG5 interacts with the retromer complex, whereas ATG5 and membrane atg8ylation affect retromer function in sorting of its cognate cargo GLUT1. Inactivation of ATG5 and of other membrane atg8ylation genes but not of the genes involved in canonical autophagy perturbs retromer-dependent trafficking. ATG5 and membrane atg8ylation display dual and combined effects on retromer function. The first one is through endolysosomal membrane maintenance ensuring proper cargo sorting by the retromer complex and its adaptor SNX27. The second one is via an association of ATG5 with the retromer complex. These and other activities of ATG5 underly the unique and often perplexing phenotypic manifestations of ATG5 observed in cells and in murine infection models 35,46,51,53–56, which served as the initial impetus for the present study. More generally, the finding that membrane atg8ylation influences retromer function further expands the range of this process specializing in homeostatic responses to membrane stress, damage and remodeling signals 25,104,122.

Mirroring the role of membrane atg8ylation on retromer function, a loss of VPS35 brings about LyHYP and influences cellular responses to lysosomal damage similarly to the inactivation of ATG5. The parallel between retromer and ATG5 is not perfect, as VPS35 knockout does not divert membrane repair protein ALIX from damaged lysosomes. This can be explained by sequestration of ALIX by the alternative conjugation complex ATG12-ATG3 formed in the absence of ATG5 35. The latter phenomenon renders ALIX unavailable for lysosomal repair 35. VPS35 knockout has no effects on ALIX, and hence the differences between ATG5 and VPS35 in the ALIX component of LyHYP. Nevertheless, inactivation of either ATG5 or VPS35 confers similarly heightened Gal3 80,81 and ubiquitin 95 responses as hallmarks of elevated lysosomal damage.

A further connectivity between the retromer system and membrane atg8ylation was observed at the level of LC3 puncta formation elicited by noncanonical triggers such as LLOMe treatment, which induces CASM and lysosomal repair. The effects of VPS35 on lysosomal functionality and morphology have been previously noted 98,99, and have been ascribed to the effects of the retromer complex on the recycling of the sorting receptors for lysosomal hydrolases, such as cation-independent mannose 6-phosphate receptor (CI-MPR) 98. We have detected significant increase in LC3 puncta in VPS35 KO cells stimulated for autophagy by starvation, consistent with the reported diminished lysosomal degradative capacity when VPS35 is absent, an effect ascribed to aberrant CI-MPR sorting 98,99. Whether CI-MPR is sorted by retromer or independently by the ESCPE-I complex 70 remains controversial 57,70,75,98,110,120,123,124. This controversy has been extended to the sorting of cation-dependent mannose 6-phosphate receptor (CD-MPR) 98,125, further confounded by the fact that retriever, which participates in receptor recycling, shares VPS29 subunit with the retromer 61. Our data showing perturbances in CI-MPR sorting in ATG5 KO and VPS35 KO cells can be best explained as indirect effects via ATG5 and VPS35 roles in endolysosomal homeostasis, which in turn may affect multiple sorting complexes.

In this study, we observed a miss-localization of the small GTPase Rab7 to lysosomes in ATG5 KO cells, paralleling the effects on GLUT1. Retromer and SNX adaptors (SNX3) interact with Rab7 108. Rab7 affects retromer-dependent sorting 106,108,109,126. Depletion of Rab7 reduces endosomal association of retromer 108 whereas retromer status affects Rab7, i.e., in cells with retromer subunits knocked out, Rab7 anomalously accumulates on lysosomes 109. Consistent with this, we observed that ATG5 KO phenocopied effects of VPS35 KO and caused Rab7 re-localization to lysosomal membranes. Rab7, like other Rabs 114, is controlled by GEFs (Mon1-Cz1), GAPs (ArmusTBC1D2A, TBC1D2B, TBC1D15), and GDI 115. It is curious that all Rab7 GAPs have LC3 interaction region (LIR) motifs and bind mATG8s: Armus/TBC12DA and TBC1D5 bind LC3A 106,113, GABARAPL1 113 and LC3C 106; TBCD12B binds all mATG8s 113,127; and TBC1D15 binds LC3A 128, LC3B, LC3C, GABARAP, GABARAP L1 and GABARAPL2 127,128. Prior elegant studies have shown that genetic ablation of the Rab7 GAP TBCD15 results in hyperactivation of Rab7 in a wrong intracellular locale (lysosome)109. In our work, carried out under basal conditions or in cells subjected to lysosomal damage, we observed both translocation of Rab7 to and entrapment of GLUT1 on lysosomes. However, we did not detect any changes in TBC1D5-mATG8s interactions, suggesting that sequestration of this Rab7 GAP by mATG8s, leading to increased plasma membrane localization of GLUT1 under glucose starvation conditions 106, is not at play under basal and endolysosmal damage or luminal alkalinization conditions tested here. We did not test glucose limitation or conditions that induce canonical autophagy or engage more complex metabolic pathways. Furthermore, we found that overexpression of constitutively active Rab7 in cells grown in glucose/nutrient-rich media causes lysosomal retention of GLUT1. This is consistent with the reports by others that perturbances in the retromer-TBC1D5 complex lead to miss-localization of Rab7 to more lysosomal-like subdomains within the endosomal system 110 and that inactivation of Rab7’s GAP TBC1D5 leads to entrapment within the endosomal system of the plasma membrane receptors as well as receptors that normally cycle back to the trans-Golgi network 129.

The role of ATG5 in retromer-dependent sorting of GLUT1 appears to have two components, the first one reflects ATG5 being a part of the E3 ligases for membrane atg8ylation and the second one reflects ATG5’s being in protein complexes with the core retromer subunits VPS26, VPS29 and VPS35. The latter association potentially explains the stronger effects of ATG5’s absence on GLUT1 sorting relative to the inactivation of other membrane atg8ylation genes. Of note, interactions between ATG5 and retromer are independent of the conjugation status of ATG5, as both ATG12—ATG5 conjugates and unconjugated ATG5 were found in protein immunoprecipitates of retromer components. This is consistent with our LC-MS/MS proteomic data indicating a conjugation status-independent increase in the proximity of ATG5 and retromer subunits during lysosomal damage. One consequence of the association between retromer and ATG5, a key component of the membrane atg8ylation apparatus, is that this could spatially direct proper membrane atg8ylation and be in part responsible for the effects of VPS35 KO on lysosomal quality and sensitivity to membrane stress or injury. In our hands, AlphaFold modeling of ATG5 with retromer components did not yield any high-confidence structures, and the best of the low-to-moderate probability structures when subjected to mutational analysis did not result in disruption of ATG5-retromer association observed by co-IPs (data not shown). Identification of the ATG5 partners that bring it to the vicinity of the retromer is yet to be accomplished and remains one of the limitations of the present study.

We used GLUT1 as a well-defined cargo sorted by the retromer in conjunction with SNX27 72,73. SNX27 is a versatile adapter, linking retromer to and controlling endosome-to-plasma membrane recycling of nearly 80 proteins such as signaling receptors, ion channels, amino acid and other nutrient transporters 73. SNX27 was also missorted in ATG5 deficient cells. This suggests that knockouts in ATG5 and in other membrane atg8ylation genes could have complex pleiotropic effects on cellular functions. Given that ATG5 has a particularly strong impact on retromer-dependent sorting, this could in part explain its unique role in the exquisite susceptibility of Atg5 mutant mice to experimental M. tuberculosis infections. Such effects likely contribute to the inflammatory action of neutrophils observed in Atg5fl/fl LysM-Cre mice 35,46,53–56. Specifically, ATG5 KO-dependent GLUT1 miss-sorting may be one of the contributors to increased M. tuberculosis pathogenesis in infection sites. In the context of tuberculosis, diabetes, which includes glucose uptake dysregulation, is associated with increased incidence of active disease and adverse outcomes 130,131. More generally, glucose uptake and metabolism are highly important for normal physiology and in various disease states including immune responses to infection, cancer, neurodegeneration, cardiovascular health, and metabolic disorders. We postulate that deficiencies or polymorphisms in membrane atg8ylation genes could have contributory roles in health and disease via GLUT1 and additional transporter proteins and signaling receptors that are dependent on retromer for their proper positioning and function in the cell.

One of the limitations of our study is that beyond the effect of membrane atg8ylation on quality of lysosomal membrane and its homeostasis, there could be more direct effects on retromer that still need to be understood. Another limitation of our study is that we have focused on basal conditions or conditions causing lysosomal damage, whereas metabolic stress including glucose excess or limitation with its multitude of metabolic effects have not been addressed. Nevertheless, we find that the membrane atg8ylation and retromer systems are intertwined, and that they affect each other’s biological outputs, one being resilience of lysosomes to basal stress or induced damage (Fig. 8G) and the other being endosomal-plasma membrane protein sorting, both being of fundamental interest and potential therapeutic value.